---
title: "Examples of how to episcout"
author: "Antonio Berlanga-Taylor"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Examples of how to use episcout}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Welcome to episcout!

This package is under development...(!)

This package that wraps up many functions that are used frequently during cleaning, exploring and visualising data sets.

It was developed using a combination of (mainly) tidyverse packages and a mixture of base R functions and data.table.

Many parameters are given as defaults out of personal preference and in order to process hundreds of thousands of observations from hundreds of variables from multiple data sets. Convenience and standardisation are higher up the agenda. This may not be your case and the choices made may not suit you. If so simply defer to your preferred R packages.

Below are a number of examples of how to use episcout functions.

All functions start with "epi\_".

Currently there are functions for:
- pre-processing: epi\_clean\_\*
- descriptive statistics: epi\_stats\_\*
- visualising: epi\_plot\_\*


## Some dummy data

```{r}
# Test set df:
n <- 20
df <- data.frame(
	var_id = rep(1:(n / 2), each = 2),
  var_to_rep = rep(c('Pre', 'Post'), n / 2),
	x = rnorm(n),
	y = rbinom(n, 1, 0.50),
	z = rpois(n, 2)
	)
df
```


**Just starting this vignette...**


## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```
