---
title: "Lower-level episcout helpers"
author: "episcout contributors"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Lower-level episcout helpers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette is a short, runnable introduction to episcout's lower-level cleaning, descriptive-statistics and plotting helpers. It uses neutral synthetic data and does not establish a scientific finding.

For a declared-dictionary EDA workflow, see the [specification-first EDA guide](specification-first-eda.html). For related restricted PostgreSQL tables, see the [longitudinal pseudonymisation guide](longitudinal-pseudonymisation.html). For explicitly declared coordinates or vector data, see the [geospatial mapping guide](geospatial-mapping-primer.html).

```{r library, message = FALSE}
library(episcout)
```

## Create neutral synthetic data

Set a seed so that the example is reproducible. The repeated `record_id` values are intentional and let the duplicate check return rows.

```{r data}
set.seed(20260812)

n <- 20
example_data <- data.frame(
  record_id = rep(seq_len(n / 2), each = 2),
  timepoint = rep(c("baseline", "follow_up"), n / 2),
  measurement = rnorm(n),
  event = rbinom(n, 1, 0.5),
  contacts = rpois(n, 2)
)

example_data
```

## Inspect duplicate values

`epi_clean_get_dups()` returns rows whose selected value occurs more than the requested frequency. Here, every `record_id` occurs twice.

```{r duplicates}
duplicates <- epi_clean_get_dups(example_data, "record_id", freq = 1)
duplicates
```

## Summarise a numeric variable

`epi_stats_numeric()` returns one row of descriptive statistics for a numeric vector. Interpret the results with the study design, data provenance and missingness rules; the helper does not make an epidemiological conclusion.

```{r numeric-summary}
summary_stats <- epi_stats_numeric(example_data$measurement)
summary_stats
```

## Plot a numeric variable

`epi_plot_hist()` creates a `ggplot2` histogram. Install the suggested `ggplot2` package to run this chunk.

```{r plotting, fig.width = 4, fig.height = 3, eval = has_ggplot2}
epi_plot_hist(example_data, "measurement")
```

```{r no-ggplot2, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("Install ggplot2 to run the plotting example.")
```
