---
title: "Longitudinal PostgreSQL pseudonymisation"
author: "episcout contributors"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Longitudinal PostgreSQL pseudonymisation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> **Technical boundary:** Pseudonymised data are not anonymous or automatically disclosure-controlled. PostgreSQL credentials, permissions, storage, backups and access controls remain external operational responsibilities.

This vignette is the detailed governance and recovery guide. For a runnable learning example using a neutral longitudinal CSV and uniquely named disposable schemas, open the installed `examples/db-to-report/walkthrough.R` script and run it section by section.

## 1. What this workflow does and when to use it

Use this workflow when related PostgreSQL tables each have one declared entity-ID column and must use the same stable pseudonymous token over time. A dictionary and linkage specification declare the requested projection and exact identity semantics, and a registry preserves token stability across runs.

The workflow performs exact matching only. It does not trim or case-fold text identifiers, infer identity from similar records, detect personal information, perform fuzzy linkage, aggregate conflicting observations or make data anonymous. PostgreSQL through RPostgres is the supported persistent backend.

Supported identity columns are PostgreSQL `text` or `varchar`, integral types and `uuid`. Fixed-width character identifiers and nondeterministic collations are rejected because they cannot guarantee this contract. Text matching distinguishes case, leading zeros and every nonblank whitespace byte; UUID matching follows PostgreSQL UUID identity rather than lexical formatting.

Use `epi_sec_pseudonym()` instead when you need a one-off in-memory bridge for one unique vector and do not need stable tokens across later calls or related tables.

## 2. PostgreSQL prerequisites

Prepare these technical conditions before R work begins:

- `source_data`, `identity_data`, `identity_registry` and `analysis_data` are distinct, existing schemas.
- Source relations and any crosswalk are ordinary PostgreSQL tables, not views or materialised views.
- The connected role has the PostgreSQL permissions required to read declared sources and crosswalks, inspect or initialise the registry, and create or replace the requested output tables.
- PostgreSQL permissions are configured outside episcout. The registry and pseudonymisation functions neither query nor change schema or table privileges.
- Named-role grants, storage encryption, backups, restore tests, retention and database logging have been reviewed outside episcout.
- The connection is not already inside a caller-managed transaction when apply begins.

episcout does not create schemas, grant application roles, manage backups or configure PostgreSQL, driver, administrator or server logging. Obtain `con` through your organisation's approved connection process; this guide deliberately contains no credential or connection-string example.

```{r inspect-connection, eval = FALSE}
# `con` is an already-open, approved RPostgres connection.
DBI::dbIsValid(con)
```

## 3. Neutral synthetic example

The example uses three neutral relations:

- `source_data.entities`: enrolment-authoritative entity metadata.
- `source_data.events`: repeated event rows.
- `source_data.measurements`: repeated measurement rows.

Each table has exactly one declared identity column. The tables may repeat the same identity; repeated identities are valid longitudinal records and are separate from duplicate-row decisions. The example refers only to synthetic structures. No observed identifiers or record-key values are copied into this guide or portable linkage metadata.

For a fresh disposable learning database whose three schemas were prepared as described above, the following unevaluated setup creates the neutral example tables and runtime-generated opaque identifiers. Do not run it against an existing or production schema.

```{r create-synthetic-tables, eval = FALSE}
DBI::dbExecute(con, "CREATE TABLE source_data.entities (entity_id text NOT NULL, entity_kind text NOT NULL)")
DBI::dbExecute(con, "CREATE TABLE source_data.events (entity_id text NOT NULL, event_sequence integer NOT NULL, event_value numeric)")
DBI::dbExecute(con, "CREATE TABLE source_data.measurements (entity_id text NOT NULL, measurement_sequence integer NOT NULL, measurement_value numeric)")

opaque_ids <- paste0("synthetic_", vapply(seq_len(2), function(index) {
  paste(sprintf("%02x", as.integer(openssl::rand_bytes(12))), collapse = "")
}, character(1)))

DBI::dbAppendTable(con, DBI::Id(schema = "source_data", table = "entities"), data.frame(
  entity_id = opaque_ids,
  entity_kind = c("kind_a", "kind_b")
))
DBI::dbAppendTable(con, DBI::Id(schema = "source_data", table = "events"), data.frame(
  entity_id = c(opaque_ids[[1]], opaque_ids[[1]], opaque_ids[[2]]),
  event_sequence = c(1L, 2L, 1L),
  event_value = c(10, 20, 30)
))
DBI::dbAppendTable(con, DBI::Id(schema = "source_data", table = "measurements"), data.frame(
  entity_id = c(opaque_ids[[1]], opaque_ids[[2]]),
  measurement_sequence = c(1L, 1L),
  measurement_value = c(100, 200)
))
rm(opaque_ids)
```

```{r inspect-source-metadata, eval = FALSE}
inventory <- epi_db_inventory(
  con,
  schema = "source_data",
  tables = c("entities", "events", "measurements"),
  row_counts = "none"
)

inventory$tables
inventory$columns
inventory$constraints
```

`row_counts = "none"` is the metadata-only default. Use estimated or exact aggregate counts only when their cost and disclosure implications have been reviewed.

## 3A. Inspect or materialise a shared identifier universe

When one declared identifier namespace is distributed across two or more source tables, define the exact observed universe from value-free technical metadata. The portable specification contains relation and column names, namespace and provenance only; it contains no identifier values.

This deliberately over-restrictive synthetic expression demonstrates error-severity findings without changing source data. A completed inspection still returns `status = "audit_complete"` and `writes = FALSE`:

```{r universe-error-audit, eval = FALSE}
universe_sources <- data.frame(
  source_schema = rep("source_data", 3),
  source_table = c("entities", "events", "measurements"),
  id_column = rep("entity_id", 3),
  identity_namespace = rep("entity_id", 3),
  provenance = rep("synthetic_metadata_definition", 3),
  stringsAsFactors = FALSE
)

over_restrictive <- epi_sec_identity_universe_spec(
  universe_sources,
  validity_regex = "^required_prefix_[0-9a-f]{24}$"
)
universe_inspection <- epi_sec_identity_universe_db(con, over_restrictive)
universe_inspection$status
universe_inspection$issues
```

The ordinary result contains only declared metadata, aggregate affected counts and fixed technical guidance. It never returns invalid values or collision members. Null, blank, regex-invalid and normalisation-collision findings have severity `error` and prevent a materialisation request from writing. Repeated valid identifiers and empty sources have severity `warning`; repeated source rows still contribute one distinct identifier to the universe.

For the runtime-generated synthetic values above, the matching expression is:

```{r universe-corrected-audit, eval = FALSE}
universe_spec <- epi_sec_identity_universe_spec(
  universe_sources,
  validity_regex = "^synthetic_[0-9a-f]{24}$"
)

universe_audit <- epi_sec_identity_universe_db(con, universe_spec)
universe_audit$source_audit
universe_audit$namespace_audit
universe_audit$overlap_audit
universe_audit$issues
```

Audit is an optional technical inspection and uses one read-only repeatable-read PostgreSQL snapshot. `n_duplicate_excess` counts valid rows beyond each source's distinct set. `n_single_source` and `n_multi_source` partition the distinct universe by how many declared relations contain an identifier. Directional coverage uses each source's distinct count as its denominator and is unavailable for an empty denominator.

A caller may request materialisation directly or after inspecting the same aggregate checks:

```{r universe-materialisation, eval = FALSE}
universe_result <- epi_sec_identity_universe_db(
  con,
  universe_spec,
  mode = "materialise",
  output_schema = "restricted_identity",
  output_table = "identity_universe",
  existing = "error"
)
```

Materialisation repeats all validation in one protected transaction and creates a new table only. The table has one unique `(identity_namespace, canonical_identifier)` row and an aggregate `source_membership_count`; it has no row-level membership table. The function uses the connected role's configured PostgreSQL permissions and does not query or change schema/table privileges. A technical error finding, existing destination or lock timeout returns `status = "not_written"`; `status = "complete"` is returned only after commit.

The result creates no token and writes no identity-registry row. A caller may declare `restricted_identity.identity_universe` as the sole `can_enrol = TRUE` source in `epi_sec_linkage_spec()`, with `canonical_identifier` as its ID column and the same namespace, for a separate registry/pseudonymisation operation. The exact universe calculation does not establish that identifiers denote distinct people.

## 4. Inventory and semantic dictionary

Create a technical and semantic dictionary from the inventory. It contains EDA labels, types, roles, units, missingness, catalogue and geographic declarations, but no privacy classification or retained-column action.

```{r dictionary-review, eval = FALSE}
dictionary <- epi_eda_dictionary_scaffold(inventory)
dictionary$provenance <- "reviewed_synthetic_metadata"

epi_eda_dictionary_validate(dictionary, catalogues = NULL)
```

For an explicitly selected categorical or binary column, profile bounded aggregate values with an exact three-key selector:

```{r catalogue-profile, eval = FALSE}
catalogue_columns <- dictionary[
  dictionary$source_column == "event_type",
  c("source_schema", "source_table", "source_column")
]

catalogue_profile <- epi_db_catalogue_profile(
  con,
  dictionary,
  columns = catalogue_columns,
  max_levels = 20
)

catalogue_profile$values
catalogue_profile$missing
```

`values` contains only non-NULL source values and counts, so its rows can be
copied into a draft normalised catalogue for human labelling and confirmation.
`missing` reports one aggregate PostgreSQL NULL count per profiled column, including zero for an empty table. PostgreSQL NULL is not a catalogue `source_value`, and this technical profile does not decide whether a sentinel represents missingness.

## 5. Linkage scaffold and specification

Create the value-free draft without reading source rows. The scaffold covers every selected semantic dictionary column in a separate `columns` component and leaves every `output_action` blank. It does not infer an ID column or decide which other columns to retain.

```{r linkage-scaffold, eval = FALSE}
draft <- epi_sec_linkage_scaffold(
  dictionary,
  tables = data.frame(
    source_schema = rep("source_data", 3),
    source_table = c("entities", "events", "measurements"),
    stringsAsFactors = FALSE
  )
)

draft
```

Complete every required field. This example uses one exact identity namespace across all three tables, allows only `entities` to enrol unseen identities and gives each output a neutral destination name. `one_row_per_entity = TRUE` is a separate row-grain declaration; it is not implied by `can_enrol = TRUE`.

```{r linkage-specification, eval = FALSE}
tables <- draft$tables
tables$id_column <- "entity_id"
tables$identity_namespace <- "entity_id"
tables$can_enrol <- tables$source_table == "entities"
tables$one_row_per_entity <- tables$source_table == "entities"
tables$destination_table <- tables$source_table
tables$provenance <- "declared_metadata"

columns <- draft$columns
id_rows <- columns$source_column == "entity_id"
columns$output_action[id_rows] <- "pseudonymise"
columns$output_action[!id_rows] <- "retain"

record_keys <- data.frame(
  source_schema = c("source_data", "source_data"),
  source_table = c("events", "measurements"),
  key_column = c("event_sequence", "measurement_sequence"),
  key_order = c(1L, 1L),
  stringsAsFactors = FALSE
)

linkage <- epi_sec_linkage_spec(
  tables = tables,
  columns = columns,
  record_keys = record_keys
)

linkage
```

The generated token is implicitly prepended to each declared record key. Do not also declare record-key columns for a table marked `one_row_per_entity = TRUE`. A table may deliberately have neither rule, but its audit can then identify exact projected duplicates only and will warn that conflicting repeated observations cannot be distinguished.

## 6. Optional database-resident crosswalk

Use a crosswalk when aliases need exact resolution. This includes reconciliation between different namespaces and deduplication aliases within the enrolment namespace. Crosswalk values stay in a caller-selected ordinary PostgreSQL table; the portable specification contains relation and column names only.

For example, if `events` uses the namespace `event_entity_id`, change that table's namespace in `tables` and describe the relation as follows. Do not place alias or canonical values in the R data frame.

```{r crosswalk-metadata, eval = FALSE}
tables$identity_namespace[tables$source_table == "events"] <- "event_entity_id"

crosswalks <- data.frame(
  crosswalk_schema = "source_data",
  crosswalk_table = "entity_crosswalk",
  alias_namespace = "event_entity_id",
  alias_id_column = "alias_id",
  canonical_namespace = "entity_id",
  canonical_id_column = "canonical_id",
  provenance = "declared_crosswalk_metadata",
  stringsAsFactors = FALSE
)

linkage <- epi_sec_linkage_spec(
  tables = tables,
  columns = columns,
  record_keys = record_keys,
  crosswalks = crosswalks
)
```

Many aliases may resolve to one entity. Conflicting mappings, chains or cycles, missing canonical targets and unmatched dependent identities produce error-severity issues that prevent apply. Unused crosswalk rows are reported by count only. episcout never infers that similar identifiers or records belong to the same entity.

## 7. Registry audit and initialisation

Start with the default read-only registry audit. It reports whether initialisation is required and never creates a table.

```{r registry-audit, eval = FALSE}
registry_audit <- epi_sec_identity_registry_init(
  con,
  registry_schema = "identity_registry"
)

registry_audit
```

When a registry is required, explicitly initialise it. The defaults use prefix `E` and 24 random bytes. Those immutable settings define one stable identity-mapping domain; reuse the registry only for sources that should share that exact mapping domain.

```{r registry-apply, eval = FALSE}
registry_apply <- epi_sec_identity_registry_init(
  con,
  registry_schema = "identity_registry",
  token_prefix = "E",
  n_bytes = 24,
  mode = "apply"
)

registry_apply
```

Apply creates or validates `registry_metadata`, `namespaces`, `entities`, `aliases`, `runs` and `run_tables` transactionally. It uses the connected role's configured PostgreSQL permissions without granting or revoking privileges. The alias table remains re-identifying plaintext. Do not edit registry rows manually; handle an incomplete registry, incompatible version or changed token setting through a separate recovery operation.

## 8. Optional audit without a data write

Run pseudonymisation in its default `mode = "audit"`. This checks schemas, inventory and drift, dictionary gates, destinations, exact identity resolution, crosswalk consistency, record keys, projected duplicates and row reconciliation without enrolling identities or creating output.

```{r workflow-audit, eval = FALSE}
audit <- epi_sec_pseudonymise_db(
  con,
  dictionary = dictionary,
  linkage = linkage,
  registry_schema = "identity_registry",
  output_schema = "analysis_data"
)

audit
audit$issues
audit$table_audit
audit$duplicate_audit
audit$manifest
```

The ordinary issue table has stable columns `issue_code`, `severity`, `stage`, `source_schema`, `source_table`, `source_column`, `n_affected`, `message` and `recommended_action`. Severity is `warning` or `error`; an error means that apply cannot produce the declared technical result. A completed audit still returns `status = "audit_complete"` and writes nothing.

Set `include_issue_values = TRUE` when identifier values for `invalid_identifier` and `unmatched_identifier` findings are needed. The returned `issue_values` component is an ordinary data frame: direct `print()` and `str()` calls display its values, and episcout does not persist it automatically. No record-key payloads, crosswalk rows or arbitrary source values are collected by this option.

## 9. Interpret statuses and resolve technical issues

The result status has one of three meanings:

| Status | Writes | Meaning | Next action |
|---|---:|---|---|
| `audit_complete` | No | The read-only inspection completed, with any error or warning issues reported separately. | Inspect the aggregate audits and issues. |
| `not_written` | No | A requested apply did not commit because a technical error issue or lock timeout was found. | Follow each error's `recommended_action`, then retry. |
| `complete` | Yes | One apply transaction committed. | Use the output and returned dictionary as requested. |

An audit-complete result says only that the inspection ran against one snapshot; it does not imply that `issues` is empty or that a later database snapshot is unchanged. Audit is optional, and apply repeats all checks in its protected transaction. Malformed arguments, unsupported types, caller-managed transactions, incompatible objects and unsafe infrastructure state remain errors rather than ordinary results.

Common error issues include missing or blank identifiers, unmatched dependent identities, conflicting or missing crosswalk targets, missing record-key values, conflicting payloads under one key, an existing destination or a lock timeout. Malformed linkage metadata are rejected earlier as specification-validation errors. Ordinary issue messages are value-free and include a recommended technical action.

## 10. Apply, verify output and confirm stable tokens

When ready to request the write, make apply explicit. A prior audit is optional because apply repeats the checks inside its protected transaction. Defaults retain exact projected duplicates and refuse existing destinations.

```{r workflow-apply, eval = FALSE}
applied <- epi_sec_pseudonymise_db(
  con,
  dictionary = dictionary,
  linkage = linkage,
  registry_schema = "identity_registry",
  output_schema = "analysis_data",
  mode = "apply",
  exact_duplicates = "report",
  existing = "error"
)

applied
applied$table_audit
applied$manifest
```

Confirm each table's `n_input`, `n_output` and `n_exact_removed` counts reconcile, and compare `duplicate_audit$n_exact_excess` with the chosen report/drop action. The `n_invalid_id`, `n_unmatched` and `n_missing_key` columns must be zero on a completed run. Inspect the destination schema's ownership and grants with the database administrator. Do not print tokens merely to verify the workflow.

To confirm stability, create a second linkage specification with new destination names, inspect or apply it, then compare token sets inside PostgreSQL and return only an aggregate Boolean. Stable tokens are guaranteed only within the same registry domain.

```{r stable-token-rerun, eval = FALSE}
repeat_tables <- linkage$tables
repeat_tables$destination_table <- paste0(repeat_tables$destination_table, "_repeat")
repeat_linkage <- epi_sec_linkage_spec(
  repeat_tables,
  columns = linkage$columns,
  record_keys = linkage$record_keys,
  crosswalks = linkage$crosswalks
)

repeat_audit <- epi_sec_pseudonymise_db(
  con, dictionary, repeat_linkage,
  registry_schema = "identity_registry",
  output_schema = "analysis_data"
)

repeat_apply <- epi_sec_pseudonymise_db(
  con, dictionary, repeat_linkage,
  registry_schema = "identity_registry",
  output_schema = "analysis_data",
  mode = "apply"
)

DBI::dbGetQuery(con, "
  SELECT NOT EXISTS (
    (SELECT entity_token FROM analysis_data.entities
     EXCEPT
     SELECT entity_token FROM analysis_data.entities_repeat)
    UNION ALL
    (SELECT entity_token FROM analysis_data.entities_repeat
     EXCEPT
     SELECT entity_token FROM analysis_data.entities)
  ) AS stable_token_set
")
```

## 11. Duplicate behaviour

Identity reuse and record duplication are different questions. Repeated identity values are valid in every table.

| Declared row rule | Observed rows | `exact_duplicates = "report"` | `exact_duplicates = "drop"` |
|---|---|---|---|
| `one_row_per_entity = TRUE` | Same token, different retained payload | Error issue; apply is not written | Error issue; apply is not written |
| Declared key `(token, key columns)` | Same key, different retained payload | Error issue; no winner is selected | Error issue; no winner is selected |
| Any rule | Entire final projection is equal using PostgreSQL null-safe equality | Count and retain every row | Count and remove only redundant projected copies |
| No declared key | Different repeated observations | Not classified as a conflict; warning explains the limitation | Not classified as a conflict; warning explains the limitation |

Record-key components must be retained, non-missing and equality-comparable. The workflow never averages, prioritises, aggregates or overwrites conflicting observations.

## 12. Hand the output dictionary to EDA

The result describes the actual pseudonymised destinations. The generated token is marked as an ID, and only catalogue definitions still referenced by retained columns are returned. The manifest identifies each `output_type` as `pseudonymised_table` without inferring a sensitivity classification.

```{r eda-handoff, eval = FALSE}
events_spec <- epi_eda_dictionary_spec(
  applied$output_dictionary,
  table = "analysis_data.events",
  catalogues = applied$output_catalogues
)

events_spec
```

Review the restricted output dictionary before EDA. Quasi-identifiers, sensitive fields, category labels and small aggregate cells may still permit disclosure. EDA does not turn pseudonymised data into anonymous or publication-safe data.

## 13. Replacement, rollback, concurrency and recovery

Apply owns one repeatable-read transaction and repeats preflight checks inside it. Registry changes, output changes and aggregate run metadata commit together; any failure rolls them all back. Transaction-scoped advisory locks are acquired in deterministic order and bounded by `lock_timeout`, which defaults to 30 seconds.

Do not wrap apply in your own transaction. After `not_written` or a sanitised apply failure, confirm `writes = FALSE`, inspect the issue table and database state, correct the problem, then start again with audit.

`existing = "replace"` is an explicit destructive choice limited to the exact declared destination ordinary tables owned by the current role. It never uses `CASCADE`; views, foreign-owned objects and dependencies prevent replacement. Prefer a new destination name when history must be retained, and coordinate replacement and recovery with the database administrator.

```{r deliberate-replacement, eval = FALSE}
replacement_audit <- epi_sec_pseudonymise_db(
  con, dictionary, linkage,
  registry_schema = "identity_registry",
  output_schema = "analysis_data",
  existing = "replace"
)

# Apply only after reviewing the audit and confirming replacement authority.
replacement_apply <- epi_sec_pseudonymise_db(
  con, dictionary, linkage,
  registry_schema = "identity_registry",
  output_schema = "analysis_data",
  mode = "apply",
  existing = "replace"
)
```

## 14. Privacy, access, backup, logging and disclosure limits

The registry is deliberately re-identifying because it stores exact source identifiers and token mappings. Restrict it more tightly than ordinary analysis output. Pseudonymised output also remains personal data because identity may be recoverable through the registry, retained fields or external information.

episcout keeps source identifiers and record-key values inside PostgreSQL by default, returns value-free ordinary audits and never writes caller-requested issue values. This package boundary does not control database queries recorded by PostgreSQL, RPostgres or infrastructure tooling; administrator access; storage snapshots; backups; replicas; monitoring; console history; or anything a caller deliberately prints or saves.

Before analysis or sharing, review least-privilege grants, retention, backup and restore controls, logging, permitted linkage purpose, quasi-identifiers, sensitive fields, small cells and the intended recipient. Pseudonymisation is not anonymisation, disclosure control or publication approval.

## 15. Troubleshooting

| Symptom | What it means | Exact next action |
|---|---|---|
| Registry audit returns `initialisation_required` | The existing schema contains no registry objects. | Run registry initialisation with `mode = "apply"` if a new registry is required. |
| Registry audit returns `incompatible` | One or more registry objects has the wrong relation kind, structure or registry version. | Inspect the registry structure and use a separate recovery operation; do not edit registry rows manually. |
| `unmatched_identifier` | A dependent identity cannot resolve exactly through its namespace, registry or declared crosswalk. | Check the namespace and crosswalk, then correct the source or metadata. Do not use fuzzy matching. |
| `conflicting_record_key` | One declared record key has different retained payloads. | Resolve the source conflict under an applicable data rule, or correct the declared key. episcout will not select a winner. |
| `destination_exists` | The output name is already occupied. | Choose a new destination, or review ownership and dependencies before explicitly auditing `existing = "replace"`. |
| `lock_timeout` error issue | Another transaction holds a required registry or destination lock. | Let the other transaction finish, verify its outcome, then audit again; do not bypass the lock. |
| Apply returns a sanitised error | PostgreSQL or the driver rejected or interrupted the operation, and the transaction rolled back. | Confirm no writes committed, inspect PostgreSQL or driver logs, correct the technical cause, then retry. |
| Audit warns that a table has no record key | Only exact projected duplicates can be assessed. | Confirm and declare the row grain and ordered key columns, or document acceptance of the limitation before apply. |
| Output is complete but sharing is planned | Completion establishes transactional workflow success only. | Perform a separate disclosure and access review; do not treat the output as anonymous. |

When seeking help, share the value-free status, issue code, relation metadata and aggregate count. Do not paste source identifiers, record-key values, tokens, credentials, connection strings or native database details into an issue or report.
