---
title: "Longitudinal PostgreSQL pseudonymisation"
author: "episcout contributors"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Longitudinal PostgreSQL pseudonymisation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> **Technical boundary:** Pseudonymised data are not anonymous or automatically disclosure-controlled. PostgreSQL credentials, permissions, storage, backups and access controls remain external operational responsibilities.

This vignette describes the technical workflow and recovery behaviour. For a runnable learning example using a neutral longitudinal CSV and uniquely named disposable schemas, open the installed `examples/db-to-report/walkthrough.R` script and run it section by section.

## 1. What this workflow does and when to use it

Use this workflow when related PostgreSQL tables each have one declared entity-ID column and must use the same pseudonymous token over time. A dictionary and linkage specification declare the requested projection and exact identity semantics. Recurring identities retain their pseudonymous identifiers when later runs reuse the same persisted registry and compatible identity mapping; separate registries do not establish cross-run stability.

The workflow performs exact matching only. It does not trim or case-fold text identifiers, infer identity from similar records, detect personal information, perform fuzzy linkage, aggregate conflicting observations or make data anonymous. PostgreSQL through RPostgres is the supported persistent backend.

Supported identity columns are PostgreSQL `text` or `varchar`, integral types and `uuid`. Fixed-width character identifiers and nondeterministic collations are rejected because they cannot guarantee this contract. Text matching distinguishes case, leading zeros and every nonblank whitespace byte; UUID matching follows PostgreSQL UUID identity rather than lexical formatting.

Use `epi_sec_pseudonym()` instead when you need a one-off in-memory bridge for one unique vector and do not need stable tokens across later calls or related tables.

`epi_sec_pseudonym()` returns a two-column tibble in input order. Setting `bridge_path` writes that re-identifying mapping as CSV; the default `overwrite = FALSE` refuses an existing path.

## 1A. Public objects and statuses

The seven public entry points return the following current objects. Their function help defines the exact formal arguments and nested data-frame columns.

| Entry point | Returned object | Modes or statuses |
|---|---|---|
| `epi_sec_pseudonym()` | A tibble with `participant_id` and `token_id` | No mode or status field |
| `epi_sec_identity_universe_spec()` | `epi_sec_identity_universe_spec`: `sources`, `normalization`, `validity_regex`, `contract_version`, `fingerprint_sha256` | Exact `identity` normalisation in contract version 2 |
| `epi_sec_identity_universe_db()` | `epi_sec_identity_universe_result`: `status`, `metadata`, `source_audit`, `namespace_audit`, `overlap_audit`, `issues` | `audit_complete`, `not_written`, `complete` |
| `epi_sec_linkage_scaffold()` | `epi_sec_linkage_scaffold`: `tables`, `columns`, `record_keys`, `crosswalks` | Editable metadata; no status field |
| `epi_sec_linkage_spec()` | `epi_sec_linkage_spec` with the same four components | Validated metadata; no status field |
| `epi_sec_identity_registry_init()` | `epi_sec_registry_result`: `status`, `mode`, `writes`, `registry_schema`, `metadata`, `objects`, `next_action` | `initialisation_required`, `incompatible`, `ready` |
| `epi_sec_pseudonymise_db()` | `epi_sec_pseudonymisation_result`: `status`, `metadata`, `identity_audit`, `table_audit`, `duplicate_audit`, `issues`, `output_dictionary`, `output_catalogues`, `manifest`, plus requested diagnostic components | `audit_complete`, `not_written`, `complete` |

A status describes only the requested technical operation. It does not decide whether processing, use, sharing or publication may proceed.

## 2. PostgreSQL prerequisites

This example uses four caller-selected schemas named `source_data`, `identity_data`, `identity_registry` and `analysis_data`. They are distinct here to make each role visible; these names and the separate `identity_data` schema are not package requirements. Create every schema named by the requested operation before calling the database functions.

- Source relations and any crosswalk are ordinary PostgreSQL tables, not views or materialised views.
- The connected role has the PostgreSQL permissions required to read declared sources and crosswalks, inspect or initialise the registry, and create or replace the requested output tables.
- PostgreSQL permissions are configured outside episcout. The registry and pseudonymisation functions neither query nor change schema or table privileges.
- Named-role grants, storage encryption, backups, restore tests, retention and database logging are configured and operated outside episcout.
- Identity-universe operations require a connection outside a caller-managed transaction. Registry and pseudonymisation apply require the same; their audit modes do not start a write transaction.

episcout does not create schemas, grant application roles, manage backups or configure PostgreSQL, driver, administrator or server logging. Supply an open `con` created through the caller's connection process; this guide deliberately contains no credential or connection-string example.

```{r inspect-connection, eval = FALSE}
# `con` is an already-open RPostgres connection.
DBI::dbIsValid(con)
```

## 3. Neutral synthetic example

The example uses three neutral relations:

- `source_data.entities`: enrolment-authoritative entity metadata.
- `source_data.events`: repeated event rows.
- `source_data.measurements`: repeated measurement rows.

Each table has exactly one declared identity column. The tables may repeat the same identity; repeated identities are valid longitudinal records and are separate from duplicate-row decisions. The example refers only to synthetic structures. No observed identifiers or record-key values are copied into this guide or portable linkage metadata.

For a fresh disposable learning database whose three schemas were prepared as described above, the following unevaluated setup creates the neutral example tables and runtime-generated opaque identifiers. Do not run it against an existing or production schema.

```{r create-synthetic-tables, eval = FALSE}
DBI::dbExecute(con, "CREATE TABLE source_data.entities (entity_id text NOT NULL, entity_kind text NOT NULL)")
DBI::dbExecute(con, "CREATE TABLE source_data.events (entity_id text NOT NULL, event_sequence integer NOT NULL, event_value numeric)")
DBI::dbExecute(con, "CREATE TABLE source_data.measurements (entity_id text NOT NULL, measurement_sequence integer NOT NULL, measurement_value numeric)")

opaque_ids <- paste0("synthetic_", vapply(seq_len(2), function(index) {
  paste(sprintf("%02x", as.integer(openssl::rand_bytes(12))), collapse = "")
}, character(1)))

DBI::dbAppendTable(con, DBI::Id(schema = "source_data", table = "entities"), data.frame(
  entity_id = opaque_ids,
  entity_kind = c("kind_a", "kind_b")
))
DBI::dbAppendTable(con, DBI::Id(schema = "source_data", table = "events"), data.frame(
  entity_id = c(opaque_ids[[1]], opaque_ids[[1]], opaque_ids[[2]]),
  event_sequence = c(1L, 2L, 1L),
  event_value = c(10, 20, 30)
))
DBI::dbAppendTable(con, DBI::Id(schema = "source_data", table = "measurements"), data.frame(
  entity_id = c(opaque_ids[[1]], opaque_ids[[2]]),
  measurement_sequence = c(1L, 1L),
  measurement_value = c(100, 200)
))
rm(opaque_ids)
```

```{r inspect-source-metadata, eval = FALSE}
inventory <- epi_db_inventory(
  con,
  schema = "source_data",
  tables = c("entities", "events", "measurements"),
  row_counts = "none"
)

inventory$tables
inventory$columns
inventory$constraints
```

`row_counts = "none"` is the metadata-only default. The caller selects estimated or exact aggregate counts when those outputs and their query cost are needed.

## 3A. Inspect or materialise a shared identifier universe

When one declared identifier namespace is distributed across two or more source tables, define the exact observed universe from value-free technical metadata. The portable specification contains relation and column names, namespace and provenance only; it contains no identifier values.

This deliberately over-restrictive synthetic expression demonstrates error-severity findings without changing source data. A completed inspection still returns `status = "audit_complete"` and `writes = FALSE`:

```{r universe-error-audit, eval = FALSE}
universe_sources <- data.frame(
  source_schema = rep("source_data", 3),
  source_table = c("entities", "events", "measurements"),
  id_column = rep("entity_id", 3),
  identity_namespace = rep("entity_id", 3),
  provenance = rep("synthetic_metadata_definition", 3),
  stringsAsFactors = FALSE
)

over_restrictive <- epi_sec_identity_universe_spec(
  universe_sources,
  validity_regex = "^required_prefix_[0-9a-f]{24}$"
)
universe_inspection <- epi_sec_identity_universe_db(con, over_restrictive)
universe_inspection$status
universe_inspection$issues
```

The ordinary result contains only declared metadata, aggregate affected counts and fixed technical guidance. It never returns invalid values or collision members. Null, blank, regex-invalid and normalisation-collision findings have severity `error` and prevent a materialisation request from writing. Repeated valid identifiers and empty sources have severity `warning`; repeated source rows still contribute one distinct identifier to the universe.

For the runtime-generated synthetic values above, the matching expression is:

```{r universe-corrected-audit, eval = FALSE}
universe_spec <- epi_sec_identity_universe_spec(
  universe_sources,
  validity_regex = "^synthetic_[0-9a-f]{24}$"
)

universe_audit <- epi_sec_identity_universe_db(con, universe_spec)
universe_audit$source_audit
universe_audit$namespace_audit
universe_audit$overlap_audit
universe_audit$issues
```

Audit is an optional technical inspection and uses one read-only repeatable-read PostgreSQL snapshot. `n_duplicate_excess` counts valid rows beyond each source's distinct set. `n_single_source` and `n_multi_source` partition the distinct universe by how many declared relations contain an identifier. Directional coverage uses each source's distinct count as its denominator and is unavailable for an empty denominator.

A caller may request materialisation directly or after inspecting the same aggregate checks:

```{r universe-materialisation, eval = FALSE}
universe_result <- epi_sec_identity_universe_db(
  con,
  universe_spec,
  mode = "materialise",
  output_schema = "identity_data",
  output_table = "identity_universe",
  existing = "error"
)
```

Materialisation repeats all validation in one protected transaction and creates a new table only. The table has one unique `(identity_namespace, canonical_identifier)` row and an aggregate `source_membership_count`; it has no row-level membership table. The function uses the connected role's configured PostgreSQL permissions and does not query or change schema/table privileges. `statement_timeout` bounds each statement and `lock_timeout` bounds the destination advisory-lock wait. A technical error finding, existing destination or lock timeout returns `status = "not_written"`; `status = "complete"` is returned only after commit. Other database failures raise a sanitised error and roll back any materialisation transaction.

The result creates no token and writes no identity-registry row. A caller may declare `identity_data.identity_universe` as the sole `can_enrol = TRUE` source in `epi_sec_linkage_spec()`, with `canonical_identifier` as its ID column and the same namespace, for a separate registry/pseudonymisation operation. The exact universe calculation does not establish that identifiers denote distinct people.

## 3B. Inspect population continuity across completed periods

Identity resolution and pseudonymisation, completed-period construction, longitudinal population/key QC, later variable-drift QC, reviewed cleaning and EDA are separate stages. After a project has constructed reviewed completed-period relations, `epi_eda_longitudinal_qc()` can compare their aggregate entity membership without collecting entity identifiers or record-key values into R.

List order is the time contract; labels are returned for interpretation but are not sorted. Every source must be an unmodified `epi_eda_postgres_source` on the same open, idle RPostgres connection. Ordinary tables, partitioned tables, views, materialised views and foreign tables are accepted through the existing source boundary.

```{r longitudinal-population-qc, eval = FALSE}
completed_periods <- list(
  period_1 = epi_eda_postgres_source(con, "analysis_data", "completed_period_1"),
  period_2 = epi_eda_postgres_source(con, "analysis_data", "completed_period_2"),
  period_3 = epi_eda_postgres_source(con, "analysis_data", "completed_period_3")
)

longitudinal_qc <- epi_eda_longitudinal_qc(
  completed_periods,
  entity_id = "token_id",
  record_key = c("token_id", "event_sequence")
)

longitudinal_qc$period_summary
longitudinal_qc$adjacent_membership
longitudinal_qc$pairwise_overlap
longitudinal_qc$history_summary
longitudinal_qc$issues
```

Null entity identifiers and blank text representations are excluded from every membership set. Repeated entity rows remain descriptive population structure. A separately declared complete simple or composite `record_key` controls only within-period uniqueness evidence; rows with any null or applicable blank key component are counted as missing keys and excluded from distinct and duplicate-key calculations.

Adjacent output reports retention, exit and entry with explicit source or target denominators. Pairwise output includes every increasing pair, including non-adjacent periods. History output groups entities by first period, last period, number of periods observed and the number of unobserved periods inside that span, making one-period entities and aggregate gaps visible without returning membership values. A zero denominator leaves its proportion unavailable rather than treating it as zero.

The four possible issues are warning-severity technical evidence for an empty period, invalid entity identifiers, missing record keys and duplicate complete record keys. Non-zero entry, exit, retention loss, reappearance and repeated entity rows do not produce warnings or scientific thresholds. Every calculation and source revalidation occurs in one read-only repeatable-read snapshot; a connection, source, catalogue, type, query, count-range or reconciliation failure returns no partial longitudinal object and leaves the caller-owned connection open and idle.

## 3C. Describe reviewed-variable continuity across completed periods

Population continuity does not describe whether reviewed variables have the same availability, storage compatibility or ordinary distribution across periods. `epi_eda_longitudinal_drift()` provides that separate descriptive evidence for an explicitly selected EDA specification. It does not identify people, return source rows, decide that a change is problematic, apply cleaning, or alter the specification.

```{r longitudinal-variable-drift, eval = FALSE}
reviewed_spec <- epi_eda_spec("reviewed-spec.csv")

longitudinal_drift <- epi_eda_longitudinal_drift(
  completed_periods,
  spec = reviewed_spec,
  variables = c("age_years", "role_group", "observation_date"),
  max_levels = 50L
)

longitudinal_drift$schema
longitudinal_drift$missingness
longitudinal_drift$missingness_adjacent
longitudinal_drift$numeric
longitudinal_drift$categorical
longitudinal_drift$temporal
```

List order defines time order, and an explicit `variables` vector retains its order. Numeric and integer fields receive canonical finite-value summaries; categorical and binary fields receive counts with total-row and observed-value denominators; dates and datetimes receive endpoint and range summaries. Every selected field receives schema and missingness evidence. Text fields are retained there and marked explicitly as unsupported for distribution summaries.

For categorical and binary fields, `max_levels` is a hard limit on declared levels, the observed domain in each period and every adjacent-period union. The default is `50L`. If any of those domains exceeds the bound, the call fails without truncating the categories or returning a partial result. Missing values appear in missingness evidence, never as an invented category. Adjacent changes are signed right-minus-left descriptive differences with their denominators retained; they are not thresholds, tests or anomaly classifications.

All validation, categorical bound checks and aggregate summaries run in one read-only repeatable-read snapshot on the caller-owned connection. The function creates no database object. A source, catalogue, query, count-range, reconciliation or transaction failure returns no partial result and leaves the connection open and reusable.

## 3D. Summarise retained-entity state transitions

Population continuity and marginal variable drift do not show how the same retained entities move between states. `epi_eda_longitudinal_transitions()` provides that separate, generic primitive for explicitly selected categorical or binary variables. It compares adjacent periods only and requires the caller to name at least one variable; selecting the entity column or any specification row whose trimmed, case-insensitive role is `id` or `identifier` fails before database work begins.

```{r longitudinal-state-transitions, eval = FALSE}
longitudinal_transitions <- epi_eda_longitudinal_transitions(
  completed_periods,
  entity_id = "token_id",
  spec = reviewed_spec,
  variables = c("role_group", "programme_status"),
  max_levels = 50L
)

longitudinal_transitions$state_audit
longitudinal_transitions$transition_summary
longitudinal_transitions$transition_counts
longitudinal_transitions$issues
```

The population contract is the same as longitudinal QC: null entity identifiers and applicable blank text representations are excluded, and repeated rows do not multiply period membership. Within one entity-period, zero distinct non-missing canonical values is a missing state, one is usable, and more than one is conflicting. Repeated copies of the same value and a mixture of one value with missing rows remain usable. A conflicting entity-period never has a state selected for it.

For each adjacent pair, `n_retained` is the distinct valid-entity intersection. It reconciles exactly to `n_eligible + n_excluded_missing + n_excluded_conflict`; conflict takes precedence when an entity is both conflicting on one side and missing on the other. `n_eligible` is the denominator for every transition proportion. When it is zero, proportions are unavailable and a `zero_eligible` warning is returned rather than a zero proportion.

An available pair uses one complete square domain. First-occurrence-deduplicated declared levels retain specification order, then unexpected usable states observed in either period follow in bytewise order. Zero-count cells remain visible, missing values never become a level, and an empty domain has no count rows. `max_levels` bounds each declaration, usable period domain and adjacent union; v1 accepts at most 50 states and therefore at most 2,500 cells per variable and pair. Exceeding any bound fails the whole call rather than truncating the domain. Counts remain exact through `2^53 - 1`, while `n_transition_cells` is an R integer.

Absent variables, incompatible storage, invalid missing-code contracts and unsupported analysis types produce period-scoped unavailable evidence. Conflicting entity-period states and zero eligible pairs are warnings. All entity-state reconciliation and transition aggregation occurs inside PostgreSQL in one read-only repeatable-read snapshot. Only explicitly selected, bounded state labels and aggregate counts leave the database; entity values, row-level entity-state records and source rows do not.

The three longitudinal interfaces answer different descriptive questions. `epi_eda_longitudinal_qc()` describes population retention, entry and exit. `epi_eda_longitudinal_drift()` describes period-level schema, missingness and marginal distributions without joining entities. `epi_eda_longitudinal_transitions()` describes adjacent state-to-state counts among retained eligible entities. None decides whether an observed change is desirable or plausible, and none derives project-specific concepts such as a job change from generic states.

## 4. Inventory and semantic dictionary

Create a technical and semantic dictionary from the inventory. It contains EDA labels, types, roles, units, missingness, catalogue and geographic declarations, but no privacy classification or retained-column action.

```{r dictionary-metadata, eval = FALSE}
dictionary <- epi_eda_dictionary_scaffold(inventory)
dictionary$analysis_type[dictionary$source_column == "entity_kind"] <- "categorical"
dictionary$provenance <- "synthetic_metadata_definition"

epi_eda_dictionary_validate(dictionary, catalogues = NULL)
```

For an explicitly selected categorical or binary column, profile bounded aggregate values with an exact three-key selector:

```{r catalogue-profile, eval = FALSE}
catalogue_columns <- dictionary[
  dictionary$source_column == "entity_kind",
  c("source_schema", "source_table", "source_column")
]

catalogue_profile <- epi_db_catalogue_profile(
  con,
  dictionary,
  columns = catalogue_columns,
  max_levels = 20
)

catalogue_profile$values
catalogue_profile$missing
```

`values` contains only non-NULL source values and counts, so its rows can be copied into a draft normalised catalogue for labelling. `missing` reports one aggregate PostgreSQL NULL count per profiled column, including zero for an empty table. PostgreSQL NULL is not a catalogue `source_value`, and this technical profile does not decide whether a sentinel represents missingness.

When a source table repeats information across two scalar columns, request that relationship explicitly rather than inferring candidates from names or inventory metadata:

```{r relationship-profile, eval = FALSE}
relationship_pairs <- data.frame(
  left_schema = "source_data",
  left_table = "entities",
  left_column = "entity_kind",
  right_schema = "source_data",
  right_table = "entities",
  right_column = "entity_kind_label"
)

relationship_profile <- epi_db_relationship_profile(
  con,
  dictionary,
  pairs = relationship_pairs,
  max_levels = 100
)

relationship_profile$summary
relationship_profile$mappings
relationship_profile$conflicts
```

This bounded profile separates SQL NULL from observed text and reports directional multiplicity without choosing a canonical field or recommending column removal. Use source documentation and review to decide whether the aggregate evidence should inform catalogue curation; it does not alter the dictionary, linkage specification or database.

## 5. Linkage scaffold and specification

Create the value-free draft without reading source rows. The scaffold covers every selected semantic dictionary column in a separate `columns` component and leaves every `output_action` blank. It does not infer an ID column or decide which other columns to retain.

```{r linkage-scaffold, eval = FALSE}
draft <- epi_sec_linkage_scaffold(
  dictionary,
  tables = data.frame(
    source_schema = rep("source_data", 3),
    source_table = c("entities", "events", "measurements"),
    stringsAsFactors = FALSE
  )
)

draft
```

Complete every required field. This example uses one exact identity namespace across all three tables, allows only `entities` to enrol unseen identities and gives each output a neutral destination name. `one_row_per_entity = TRUE` is a separate row-grain declaration; it is not implied by `can_enrol = TRUE`.

```{r linkage-specification, eval = FALSE}
tables <- draft$tables
tables$id_column <- "entity_id"
tables$identity_namespace <- "entity_id"
tables$can_enrol <- tables$source_table == "entities"
tables$one_row_per_entity <- tables$source_table == "entities"
tables$destination_table <- tables$source_table
tables$provenance <- "declared_metadata"

columns <- draft$columns
id_rows <- columns$source_column == "entity_id"
columns$output_action[id_rows] <- "pseudonymise"
columns$output_action[!id_rows] <- "retain"

record_keys <- data.frame(
  source_schema = c("source_data", "source_data"),
  source_table = c("events", "measurements"),
  key_column = c("event_sequence", "measurement_sequence"),
  key_order = c(1L, 1L),
  stringsAsFactors = FALSE
)

linkage <- epi_sec_linkage_spec(
  tables = tables,
  columns = columns,
  record_keys = record_keys
)

linkage
```

The generated token is implicitly prepended to each declared record key. Do not also declare record-key columns for a table marked `one_row_per_entity = TRUE`. A table may deliberately have neither rule, but its audit can then identify exact projected duplicates only and will warn that conflicting repeated observations cannot be distinguished.

## 6. Optional database-resident crosswalk

Use a crosswalk when aliases need exact resolution. This includes reconciliation between different namespaces and deduplication aliases within the enrolment namespace. Crosswalk values stay in a caller-selected ordinary PostgreSQL table; the portable specification contains relation and column names only.

For example, if `events` uses the namespace `event_entity_id`, change that table's namespace in `tables` and describe the relation as follows. Do not place alias or canonical values in the R data frame.

```{r crosswalk-metadata, eval = FALSE}
tables$identity_namespace[tables$source_table == "events"] <- "event_entity_id"

crosswalks <- data.frame(
  crosswalk_schema = "identity_data",
  crosswalk_table = "entity_crosswalk",
  alias_namespace = "event_entity_id",
  alias_id_column = "alias_id",
  canonical_namespace = "entity_id",
  canonical_id_column = "canonical_id",
  provenance = "declared_crosswalk_metadata",
  stringsAsFactors = FALSE
)

linkage <- epi_sec_linkage_spec(
  tables = tables,
  columns = columns,
  record_keys = record_keys,
  crosswalks = crosswalks
)
```

Many aliases may resolve to one entity. Conflicting mappings, chains or cycles, missing canonical targets and unmatched dependent identities produce error-severity issues that prevent apply. Unused crosswalk rows are reported by count only. episcout never infers that similar identifiers or records belong to the same entity.

### Auditing and materialising an exact alias crosswalk

When the source itself explicitly co-observes two identifier systems, use
`epi_sec_alias_spec()` and `epi_sec_alias_db()` rather than constructing an
unreviewed crosswalk externally. The specification names the source relation,
the two identifier columns, and the caller-selected alias and canonical
namespaces; it contains no identifier values. An audit returns aggregate,
value-free counts for incomplete identifiers, duplicate pairs and mapping
multiplicity. Alias-to-many-canonical and canonical-to-many-alias evidence is
an error, not a precedence decision.

```{r alias-crosswalk, eval = FALSE}
alias_spec <- epi_sec_alias_spec(
  data.frame(
    source_schema = "identity_data",
    source_table = "explicit_identifier_pairs",
    alias_id_column = "secondary_identifier",
    canonical_id_column = "primary_identifier",
    alias_namespace = "secondary_namespace",
    canonical_namespace = "primary_namespace",
    provenance = "declared_exact_co_observation",
    stringsAsFactors = FALSE
  )
)
alias_audit <- epi_sec_alias_db(con, alias_spec)
```

After all error findings are resolved, materialisation creates a new ordinary
PostgreSQL table with `alias_id` and `canonical_id`; it never replaces an
existing table. Pass that relation and its two column names to
`epi_sec_linkage_spec()` as normal crosswalk metadata. This establishes only
the exact relationships the caller declared. A fallback identifier choice is
not alias resolution, and neither mechanism infers identity from similar
records.

## 7. Registry audit and initialisation

Start with the default read-only registry audit. It reports whether initialisation is required and never creates a table.

```{r registry-audit, eval = FALSE}
registry_audit <- epi_sec_identity_registry_init(
  con,
  registry_schema = "identity_registry"
)

registry_audit
```

To create a registry in an empty schema, explicitly request initialisation. The defaults use prefix `E` and 24 random bytes. Those token settings are immutable for the registry. Assignments persist across later calls to the same registry, while identity namespaces and crosswalks determine which source identifiers resolve to the same token. A separately initialised registry has independent assignments and cannot be used as evidence of cross-run identity stability.

```{r registry-apply, eval = FALSE}
registry_apply <- epi_sec_identity_registry_init(
  con,
  registry_schema = "identity_registry",
  token_prefix = "E",
  n_bytes = 24,
  mode = "apply"
)

registry_apply
```

Apply creates or validates `registry_metadata`, `namespaces`, `entities`, `aliases`, `runs` and `run_tables` transactionally. It uses the connected role's configured PostgreSQL permissions without granting or revoking privileges. The alias table remains re-identifying plaintext. Do not edit registry rows manually; handle an incomplete registry, incompatible version or changed token setting through a separate recovery operation.

## 8. Optional audit without a data write

Run pseudonymisation in its default `mode = "audit"`. This checks schemas, inventory and drift, dictionary gates, destinations, exact identity resolution, crosswalk consistency, record keys, projected duplicates and row reconciliation without enrolling identities or creating output.

```{r workflow-audit, eval = FALSE}
audit <- epi_sec_pseudonymise_db(
  con,
  dictionary = dictionary,
  linkage = linkage,
  registry_schema = "identity_registry",
  output_schema = "analysis_data"
)

audit
audit$issues
audit$table_audit
audit$duplicate_audit
audit$manifest
```

The ordinary issue table has stable columns `issue_code`, `severity`, `stage`, `source_schema`, `source_table`, `source_column`, `n_affected`, `message` and `recommended_action`. Severity is `warning` or `error`; an error means that apply cannot produce the declared technical result. A completed audit still returns `status = "audit_complete"` and writes nothing.

Set `include_issue_values = TRUE` when identifier values for `invalid_identifier` and `unmatched_identifier` findings are needed. The returned `issue_values` component is an ordinary data frame: direct `print()` and `str()` calls display its values, and episcout does not persist it automatically. No record-key payloads, crosswalk rows or arbitrary source values are collected by this option.

## 9. Interpret statuses and resolve technical issues

The result status has one of three meanings:

| Status | Writes | Meaning | Next action |
|---|---:|---|---|
| `audit_complete` | No | The read-only inspection completed, with any error or warning issues reported separately. | Inspect the aggregate audits and issues. |
| `not_written` | No | A requested apply did not commit because a technical error issue or lock timeout was found. | Follow each error's `recommended_action`, then retry. |
| `complete` | Yes | One apply transaction committed. | Use the output and returned dictionary as requested. |

An audit-complete result says only that the inspection queries completed; it does not imply that `issues` is empty or that concurrent or later database state is unchanged. Pseudonymisation audit does not acquire the apply locks or own one repeatable-read snapshot. Audit is optional, and apply repeats all checks in its protected transaction. Malformed arguments, unsupported types, caller-managed apply transactions, incompatible objects and unsafe infrastructure state remain errors rather than ordinary results.

Common error issues include missing or blank identifiers, unmatched dependent identities, conflicting or missing crosswalk targets, missing record-key values, conflicting payloads under one key, an existing destination or a lock timeout. Malformed linkage metadata are rejected earlier as specification-validation errors. Ordinary issue messages are value-free and include a recommended technical action.

## 10. Apply, verify output and confirm stable tokens

To request the write, make apply explicit. A prior audit is optional because apply repeats the checks inside its protected transaction. Defaults retain exact projected duplicates and refuse existing destinations.

```{r workflow-apply, eval = FALSE}
applied <- epi_sec_pseudonymise_db(
  con,
  dictionary = dictionary,
  linkage = linkage,
  registry_schema = "identity_registry",
  output_schema = "analysis_data",
  mode = "apply",
  exact_duplicates = "report",
  existing = "error"
)

applied
applied$table_audit
applied$manifest
```

Confirm each table's `n_input`, `n_output` and `n_exact_removed` counts reconcile, and compare `duplicate_audit$n_exact_excess` with the chosen report/drop action. The `n_invalid_id`, `n_unmatched` and `n_missing_key` columns are zero on a completed run. The function neither queries nor changes schema or table privileges; PostgreSQL applies the connected role's ownership and default-privilege configuration to newly created tables. Do not print tokens merely to verify the workflow.

To confirm stability, create a second linkage specification with new destination names, inspect or apply it against the same persisted registry, then compare token sets inside PostgreSQL and return only an aggregate Boolean. Stability applies only to that registry assignment: registry, identity namespace and any alias/crosswalk mapping must resolve the identifier to the same stored entity token. A separate registry is not a stability comparison.

```{r stable-token-rerun, eval = FALSE}
repeat_tables <- linkage$tables
repeat_tables$destination_table <- paste0(repeat_tables$destination_table, "_repeat")
repeat_linkage <- epi_sec_linkage_spec(
  repeat_tables,
  columns = linkage$columns,
  record_keys = linkage$record_keys,
  crosswalks = linkage$crosswalks
)

repeat_audit <- epi_sec_pseudonymise_db(
  con, dictionary, repeat_linkage,
  registry_schema = "identity_registry",
  output_schema = "analysis_data"
)

repeat_apply <- epi_sec_pseudonymise_db(
  con, dictionary, repeat_linkage,
  registry_schema = "identity_registry",
  output_schema = "analysis_data",
  mode = "apply"
)

DBI::dbGetQuery(con, "
  SELECT NOT EXISTS (
    (SELECT entity_token FROM analysis_data.entities
     EXCEPT
     SELECT entity_token FROM analysis_data.entities_repeat)
    UNION ALL
    (SELECT entity_token FROM analysis_data.entities_repeat
     EXCEPT
     SELECT entity_token FROM analysis_data.entities)
  ) AS stable_token_set
")
```

## 11. Duplicate behaviour

Identity reuse and record duplication are different questions. Repeated identity values are valid in every table.

| Declared row rule | Observed rows | `exact_duplicates = "report"` | `exact_duplicates = "drop"` |
|---|---|---|---|
| `one_row_per_entity = TRUE` | Same token, different retained payload | Error issue; apply is not written | Error issue; apply is not written |
| Declared key `(token, key columns)` | Same key, different retained payload | Error issue; no winner is selected | Error issue; no winner is selected |
| Any rule | Entire final projection is equal using PostgreSQL null-safe equality | Count and retain every row | Count and remove only redundant projected copies |
| No declared key | Different repeated observations | Not classified as a conflict; warning explains the limitation | Not classified as a conflict; warning explains the limitation |

Record-key components must be retained, non-missing and equality-comparable. The workflow never averages, prioritises, aggregates or overwrites conflicting observations.

## 12. Hand the output dictionary to EDA

The result describes the actual pseudonymised destinations. The generated token is marked as an ID, and only catalogue definitions still referenced by retained columns are returned. The manifest identifies each `output_type` as `pseudonymised_table` without inferring a sensitivity classification.

```{r eda-handoff, eval = FALSE}
events_spec <- epi_eda_dictionary_spec(
  applied$output_dictionary,
  table = "analysis_data.events",
  catalogues = applied$output_catalogues
)

events_spec
```

The output dictionary describes the generated token and retained columns; it does not classify their permitted use. Retained fields, category labels and small aggregate cells can still contribute to re-identification or disclosure. EDA does not turn pseudonymised data into anonymous or automatically disclosure-controlled data.

## 13. Replacement, rollback, concurrency and recovery

Apply acquires bounded session advisory locks in deterministic order, opens one repeatable-read transaction and transfers protection to transaction-scoped advisory locks before releasing the session locks. It then repeats the checks inside the transaction. Registry changes, output changes and aggregate run metadata commit together; any failure rolls them all back. `lock_timeout` bounds the initial lock wait and defaults to 30 seconds; transaction-scoped locks are released by commit or rollback.

Do not wrap apply in your own transaction. A `not_written` result has `metadata$writes = FALSE`; inspect its issue table and the observed database state, correct the technical cause and retry audit or apply. A sanitised apply error has no result object. Errors before the protected transaction occur before the write phase; errors after it begins roll that transaction back. Inspect PostgreSQL state and available driver or server logs, correct the cause and retry from an idle connection.

`existing = "replace"` is an explicit destructive choice limited to the exact declared destination ordinary tables owned by the current role. It never uses `CASCADE`; views, foreign-owned objects and dependencies prevent replacement. Use a new destination name when the existing table must be retained. Database administration and recovery remain outside episcout.

```{r deliberate-replacement, eval = FALSE}
replacement_audit <- epi_sec_pseudonymise_db(
  con, dictionary, linkage,
  registry_schema = "identity_registry",
  output_schema = "analysis_data",
  existing = "replace"
)

# Apply repeats the checks and requests replacement of the declared tables.
replacement_apply <- epi_sec_pseudonymise_db(
  con, dictionary, linkage,
  registry_schema = "identity_registry",
  output_schema = "analysis_data",
  mode = "apply",
  existing = "replace"
)
```

The preliminary replacement audit is optional. `existing = "replace"` is the caller's explicit operation selection, not a permission decision made by episcout.

## 14. Reversibility and external operational boundaries

The registry is re-identifying because it stores exact source identifiers and token mappings. Pseudonymised output may also remain identifiable through the registry, retained fields or external information. episcout does not control who can access either database location.

episcout keeps source identifiers and record-key values inside PostgreSQL by default, returns value-free ordinary audits and does not persist caller-requested `issue_values`. This package boundary does not control database queries recorded by PostgreSQL, RPostgres or infrastructure tooling; administrator access; storage snapshots; backups; replicas; monitoring; console history; or anything a caller deliberately prints or saves.

PostgreSQL grants, retention, backup and restore controls, logging and decisions about downstream analysis, sharing or publication remain external responsibilities. Pseudonymisation is not anonymisation or automatic disclosure control, and no episcout status supplies publication or disclosure permission.

## 15. Troubleshooting

| Symptom | What it means | Exact next action |
|---|---|---|
| Registry audit returns `initialisation_required` | The existing schema contains no registry objects. | Run registry initialisation with `mode = "apply"` if a new registry is required. |
| Registry audit returns `incompatible` | One or more registry objects has the wrong relation kind, structure or registry version. | Inspect the registry structure and use a separate recovery operation; do not edit registry rows manually. |
| `unmatched_identifier` | A dependent identity cannot resolve exactly through its namespace, registry or declared crosswalk. | Check the namespace and crosswalk, then correct the source or metadata. Do not use fuzzy matching. |
| `conflicting_record_key` | One declared record key has different retained payloads. | Resolve the source conflict under an applicable data rule, or correct the declared key. episcout will not select a winner. |
| `destination_exists` | The output name is already occupied. | Choose a new destination, or inspect PostgreSQL ownership and dependencies before explicitly selecting `existing = "replace"`. |
| `lock_timeout` error issue | Another transaction holds a required registry or destination lock. | Let the other transaction finish, verify its outcome, then retry audit or apply; do not bypass the lock. |
| Apply raises a sanitised error | PostgreSQL or the driver rejected or interrupted the operation; no result object is returned. | If the protected transaction began, it rolled back; preflight errors occurred before writes. Inspect database state and available logs, correct the cause and retry. |
| Audit warns that a table has no record key | Only exact projected duplicates can be assessed. | Declare the row grain and ordered key columns when conflict detection is needed; otherwise the limitation remains visible in the result. |
| A completed output is being considered for downstream use | Completion establishes transactional workflow success only. | Decide use, access, sharing and publication outside episcout; do not treat the output as anonymous. |

When seeking help, share the value-free status, issue code, relation metadata and aggregate count. Do not paste source identifiers, record-key values, tokens, credentials, connection strings or native database details into an issue or report.
