---
title: "Specification-first exploratory data analysis"
author: "Antonio Berlanga-Taylor"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Specification-first exploratory data analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = has_ggplot2)
```

```{r eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("This vignette needs ggplot2. Install it to run the examples.")
```

Specification-first EDA uses an analyst-declared semantic dictionary for schema checks, missingness, summaries, plots, optional stratification and optional point maps. The same declaration can drive an in-memory or PostgreSQL workflow.

episcout creates the outputs explicitly requested by the analyst and does not decide whether they may be shared.

## Generate a lean semantic dictionary

When data arrive without a dictionary, `epi_eda_spec_scaffold()` uses storage metadata only. Its exact fields are `name`, `label`, `type`, `role`, `units`, `levels`, `min`, `max`, `missing_codes`, `required`, `group`, `description`, `geo_role`, `geo_pair` and `geo_crs`. It does not include observed counts, candidate declarations, approval states or privacy policy.

```{r}
library(episcout)

received_data <- data.frame(
  age = c(42, 57, 61, NA),
  study_group = c("A", "B", "A", "B"),
  longitude = c(-3, -1, 1, 3),
  latitude = c(51, 52, 53, 54),
  stringsAsFactors = FALSE
)

spec <- epi_eda_spec_scaffold(received_data)
names(spec)
```

Storage-derived types are a starting point. Declare the intended categorical semantics and the coordinate pair explicitly; episcout does not infer either from names or roles.

```{r}
spec$role <- c("covariate", "exposure", "coordinate", "coordinate")
spec$required <- TRUE
spec$type[spec$name == "study_group"] <- "categorical"
spec$levels[spec$name == "study_group"] <- "A;B"
spec$geo_role[spec$name %in% c("longitude", "latitude")] <- c("x", "y")
spec$geo_pair[spec$name %in% c("longitude", "latitude")] <- "site"
spec$geo_crs[spec$name %in% c("longitude", "latitude")] <- "4326"
spec <- epi_eda_spec(spec)
```

`missing_codes` accepts semicolon-separated declared sentinels. Preparation is all-or-nothing: incompatible storage, unsafe conversion or unexpected categorical values produce factual audit findings rather than partial transformations. Roles describe variables; `id`, `identifier` and coordinate roles do not suppress ordinary summaries, plots or strata.

```{r}
assessment <- epi_eda_prepare(received_data, spec, mode = "audit")
assessment$audit

prepared <- epi_eda_prepare(received_data, spec, mode = "apply")
prepared$metadata
```

## Use the intake workflow

With `spec = NULL`, intake generates the lean dictionary, saves it as `specification.csv`, uses it, and continues. Its processing statuses are `complete`, `audit_complete` and `blocked`.

```{r}
generated_dir <- tempfile("episcout-generated-intake-")
generated <- epi_eda_intake_run(
  received_data,
  output_dir = generated_dir,
  render = FALSE
)
generated$status
generated$spec$data
generated$manifest
```

Supply a semantic dictionary when declarations differ from storage or when preparation is required.

```{r}
intake_dir <- tempfile("episcout-complete-intake-")
intake <- epi_eda_intake_run(
  received_data,
  spec,
  output_dir = intake_dir,
  prepare = "apply",
  strata = "study_group",
  render = FALSE
)
intake$status
intake$stratified$groups
intake$categorical_display
intake$table1
```

Core manifests have exactly `artifact`, `type`, `path`, `status` and `checksum_md5`. Intake never writes source or prepared rows.

## Request declared point maps

Mapping is off by default. `maps = TRUE` requests one geometry-only map per declared pair; each unique `map_vars` entry requests one additional thematic map per pair. Themes may be numeric, integer, categorical, binary or text. Date and datetime themes are unsupported. No thematic variable is selected automatically.

```{r, fig.width = 6, fig.height = 4}
mapped <- epi_eda_run(
  received_data,
  spec,
  maps = TRUE,
  map_vars = "study_group",
  max_map_points = 10000L
)
mapped$geo
mapped$map_inventory
mapped$maps[["map-p001-geometry"]]
mapped$maps[["map-p001-v002"]]
```

Maps are created only when every row in a pair is complete, finite, within EPSG:4326 ranges when applicable, and the source has at most `max_map_points` rows. A failed pair records `not_created`, preserves the rest of EDA, and is never partially mapped or truncated. Declared thematic missing codes use the existing missing-value colour. Integrated EDA creates coordinate-derived points only; use explicit `epi_geo_*` functions for existing `sf`, line, polygon, boundary, basemap or PostGIS geometry workflows.

The same options are available on `epi_eda_intake_run()` and `epi_eda_render_report()`. Bundle-producing runs write deterministic SVGs under `maps/`; HTML reports embed their maps and inventory.

## Summaries and Table 1

The canonical summary contains `variables`, `numeric`, `categorical`, `text`, `temporal` and `skipped`. Every specification row remains accounted for. Numeric summaries separate finite values from infinities, categorical summaries retain declared zero-count levels and temporal summaries state their units.

```{r}
results <- epi_eda_run(prepared$data, spec)
results$summaries$variables

stratified <- epi_eda_profile_stratified(
  prepared$data,
  spec,
  strata = "study_group"
)
epi_eda_table1(stratified)
```

`epi_eda_categorical_display()` is the shared aggregate calculation behind categorical presentation. It combines standard and declared-code missing values in one explicit missing row and retains zero denominators as integer zero with `NA_real_` proportion.

```{r}
compatibility <- epi_eda_categorical_display(stratified)
column <- epi_eda_categorical_display(stratified, basis = "column")
row <- epi_eda_categorical_display(stratified, basis = "row")
overall <- epi_eda_categorical_display(stratified, basis = "overall")

epi_eda_table1(stratified, basis = "column")
```

The default `compatibility` basis uses observed non-missing values for ordinary levels and all group rows for the missing level. `column` uses all rows within each group, `row` distributes each level across non-Overall strata, and `overall` uses the included analysis population. Row percentages require a stratified result. `epi_eda_run()` and intake publish the compatibility table as `categorical_display.csv`; the HTML reports show the same numeric fields.

## PostgreSQL-backed EDA

The caller owns an open idle RPostgres connection and supplies schema and relation identifiers separately. Arbitrary SQL, dotted relation names, generic DBI connections and caller-managed transactions are not accepted.

```{r eval = FALSE}
con <- DBI::dbConnect(
  RPostgres::Postgres(),
  host = Sys.getenv("PGHOST"),
  dbname = Sys.getenv("PGDATABASE"),
  user = Sys.getenv("PGUSER")
)
source <- epi_eda_postgres_source(con, "eda_fixture", "observations")

db_results <- epi_eda_db_run(
  source,
  spec,
  output_dir = "20260809_eda_cycle",
  plots = TRUE,
  maps = TRUE,
  map_vars = "study_group",
  max_map_points = 10000L,
  layout = "delivery"
)
DBI::dbDisconnect(con)
```

All profiling reads and the map bound check occur in one read-only repeatable-read snapshot. Coordinates and thematic observations are collected only for requested, map-ready, within-limit pairs. The query projects only ready-pair coordinates and explicitly selected themes, uses a defensive `max_map_points + 1` bound, and fails rather than truncates or silently drops rows. Rendering and atomic filesystem publication happen after the snapshot ends.

`layout = "delivery"` makes `README.md` and `reports/eda-report.html` the human entry points. `QA_QC/` retains aggregate checks and summaries, `plot_data/` retains compact aggregate plot inputs but never map coordinates/themes or source rows, `plots/` and `maps/` retain deterministic SVGs, and `run_manifests/` retains provenance, timings and the five-column checksum manifest. Frequency companions use `compact-plot-data-2`: their established plot columns come first, followed by the shared categorical display fields. Valid legacy four-column `compact-plot-data-1` companions are reconstructed from validated summary aggregates in memory and are not rewritten. The caller supplies the exact root; a `YYYYMMDD_eda_cycle` name is recommended rather than inferred. The default `layout = "bundle"` preserves the flat bundle and its metadata schemas.

Existing flat and delivery bundles can be rendered without a live connection. The renderer re-reads and validates the on-disk bundle; a result object contributes only its output directory.

```{r eval = FALSE}
report <- epi_eda_render_db_report("postgres-eda-bundle")
# Use overwrite = TRUE only to replace an already owned README/report.
```

The extended PostgreSQL dictionary is technical and semantic. Catalogue profiling uses an explicit three-key selector:

```{r eval = FALSE}
dictionary <- epi_eda_dictionary_scaffold(epi_db_inventory(con, "study"))
columns <- dictionary[
  dictionary$source_column == "study_group",
  c("source_schema", "source_table", "source_column")
]
profile <- epi_db_catalogue_profile(con, dictionary, columns, max_levels = 50L)
```

Specialised privacy classification and retained-column actions belong to `epi_sec_linkage_scaffold()` and `epi_sec_linkage_spec(columns = ...)`, not the EDA dictionary. A pseudonymisation output dictionary is semantic and can pass directly to `epi_eda_dictionary_spec()`.

## Breaking-change migration

This release intentionally rejects old contracts:

- Replace scaffold evidence/candidate/review fields with the 15 semantic fields shown above; regenerate with `epi_eda_spec_scaffold()`.
- Remove `privacy_class`, `analytic_action`, `validation_status` and `profile_catalogue` from extended EDA dictionaries. Put security column policy in `linkage$columns`.
- Remove ordinary catalogue `validation_status`; preserve labels, missingness, ordering and provenance.
- Pass an exact `source_schema`, `source_table`, `source_column` selector to `epi_db_catalogue_profile()`.
- Replace core six-column sensitivity manifests with the five-column manifest contract.
- Regenerate old three-component linkage objects with `tables`, `columns`, `record_keys` and `crosswalks`.
- Replace database-to-report examples that extract rows with `epi_eda_db_run(layout = "delivery")` or render an existing aggregate bundle with `epi_eda_render_db_report()`. Direct in-memory reporting remains available through `epi_eda_render_report()`.

Old objects fail with concise migration guidance; no warning-only compatibility shim is provided.

## Write in-memory outputs and render HTML

```{r eval = FALSE}
output_dir <- file.path(tempdir(), "episcout-eda")
dir.create(output_dir, showWarnings = FALSE)
epi_eda_render_report(
  received_data,
  spec,
  output_dir,
  maps = TRUE,
  map_vars = "study_group"
)
```

For a file-based project, use `epi_eda_create_project("my-eda-project")`. The lower-level `epi_eda_*` functions remain available for ordinary scripts.
