---
title: "Specification-first exploratory data analysis"
author: "Antonio Berlanga-Taylor"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Specification-first exploratory data analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = has_ggplot2)
```

```{r eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("This vignette needs ggplot2. Install it to run the examples.")
```

Specification-first EDA starts with a reviewed semantic dictionary, then uses that dictionary for preparation, schema and missingness checks, summaries, plots, optional stratification, optional point maps and report delivery. The data-frame and PostgreSQL entry points share the same EDA specification and aggregate summary contracts, but their preparation and output ownership differ.

episcout creates the outputs explicitly requested by the analyst and does not decide whether they may be shared. Aggregate outputs can still reveal sensitive names, categories, small cells or point locations, so review the complete output root before delivery.

## 1. Choose the data source

Use the data-frame path when the caller has already loaded an authorised in-memory data frame. `epi_eda_intake_run()` is the guided audit/apply and owned-bundle entry point. `epi_eda_run()` and `epi_eda_render_report()` are lower-level analysis and reporting entry points; they do not call `epi_eda_prepare()`.

When a caller supplies a synthetic fixture, set `data_origin = "caller_declared_synthetic"` so workflow metadata and reports do not mislabel it as observed. This declaration is explicitly unverified: synthetic provenance and fixture integrity do not establish anonymisation, disclosure safety, representativeness, scientific validity or production readiness. `epi_eda_inject_missingness()` can add exact seeded counts of R missing values and literal blanks to a neutral fixture without changing the complete-data default of `epi_eda_generate_synthetic_data()`.

Use the PostgreSQL path when observations must remain in one supported database relation. The caller creates and owns the connection and credentials, then passes separate schema and relation identifiers to `epi_eda_postgres_source()`. episcout does not create a server, manage credentials, accept arbitrary SQL or prepare database values.

The installed `project-template/README.md` provides a file-based data-frame scaffold. The installed `examples/db-to-report/README.md` and `walkthrough.R` provide a runnable synthetic example of database inventory, separate pseudonymisation policy and PostgreSQL EDA delivery.

## 2. Create and review the dictionary

For a data frame without a dictionary, `epi_eda_spec_scaffold()` uses storage metadata only. It returns the 16 semantic fields `name`, `label`, `database_type`, `analysis_type`, `role`, `units`, `levels`, `min`, `max`, `missing_codes`, `required`, `group`, `description`, `geo_role`, `geo_pair` and `geo_crs`. `database_type` records the storage family and `analysis_type` selects the R/EDA treatment. It does not infer scientific roles, sentinel missing values, units, validation ranges, geographic meaning, privacy policy or approval status, and it does not inspect observed values for evidence.

```{r}
library(episcout)

received_data <- data.frame(
  age = c(42L, 57L, 61L, NA_integer_),
  study_group = c("A", "B", "A", "B"),
  longitude = c(-3, -1, 1, 3),
  latitude = c(51, 52, 53, 54),
  stringsAsFactors = FALSE
)

spec <- epi_eda_spec_scaffold(received_data)
names(spec)
```

Storage-derived types are only a starting point. Declare the intended semantics and review them against source documentation. In this neutral example, the analyst declares a categorical exposure and one coordinate pair explicitly; episcout does not infer them from names or roles.

```{r}
spec$role <- c("covariate", "exposure", "coordinate", "coordinate")
spec$required <- TRUE
spec$analysis_type[spec$name == "study_group"] <- "categorical"
spec$levels[spec$name == "study_group"] <- "A;B"
spec$geo_role[spec$name %in% c("longitude", "latitude")] <- c("x", "y")
spec$geo_pair[spec$name %in% c("longitude", "latitude")] <- "site"
spec$geo_crs[spec$name %in% c("longitude", "latitude")] <- "4326"
spec <- epi_eda_spec(spec)
```

`missing_codes` accepts semicolon-separated declared sentinels. Standard missing values and these declared codes are treated as missing in EDA calculations. Roles describe variables; `id`, `identifier` and coordinate roles do not suppress ordinary summaries, plots or strata.

For PostgreSQL, inventory the caller-selected schema, then create the extended technical and semantic dictionary. Inventory reads metadata by default; request `row_counts = "exact"` only when the aggregate count is needed and acceptable for the source. Convert one reviewed active table into the same 16-field EDA specification.

```{r eval = FALSE}
inventory <- epi_db_inventory(
  con,
  schema = "study",
  tables = "observations",
  row_counts = "none"
)
dictionary <- epi_eda_dictionary_scaffold(inventory)

# Edit semantic fields and optional catalogue definitions, then validate them.
epi_eda_dictionary_validate(dictionary, catalogues)
spec <- epi_eda_dictionary_spec(
  dictionary,
  table = "study.observations",
  catalogues = catalogues
)
```

The extended dictionary contains technical source metadata and semantic declarations. `epi_db_catalogue_profile()` reads values only for an explicit `source_schema`, `source_table`, `source_column` selection and a caller-supplied `max_levels` bound. Privacy classification, column retention and pseudonymisation policy do not belong in either EDA dictionary; use the separate `epi_sec_*` workflow when those controls are required. A pseudonymisation result's semantic `output_dictionary` and `output_catalogues` can pass directly to `epi_eda_dictionary_spec()`.

Before deciding how repeated code and label fields should be represented in a reviewed specification, inspect an explicit same-table pair with a separate combination bound. The profiler returns aggregate mappings, missingness and multiplicity exceptions; it does not decide that fields are equivalent, redundant or canonical.

```{r relationship-profile, eval = FALSE}
relationship_pairs <- data.frame(
  left_schema = "study",
  left_table = "observations",
  left_column = "group_code",
  right_schema = "study",
  right_table = "observations",
  right_column = "group_label"
)

relationship_profile <- epi_db_relationship_profile(
  con,
  dictionary,
  pairs = relationship_pairs,
  max_levels = 100
)

relationship_profile$summary
relationship_profile$mappings
relationship_profile$conflicts

# Only after reviewing the relationship evidence and source documentation,
# profile the selected catalogue field and author its semantics explicitly.
catalogue_profile <- epi_db_catalogue_profile(
  con,
  dictionary,
  columns = data.frame(
    source_schema = "study",
    source_table = "observations",
    source_column = "group_code"
  ),
  max_levels = 100
)
catalogue_profile$values
catalogue_profile$missing

catalogues <- data.frame(
  catalog_name = c("study_group", "study_group"),
  source_value = c("A", "B"),
  label = c("Group A", "Group B"),
  display_order = c(1, 2),
  is_missing = c(FALSE, FALSE),
  provenance = c("reviewed source definition", "reviewed source definition")
)
dictionary$analysis_type[dictionary$source_column == "group_code"] <-
  "categorical"
dictionary$catalog_name[dictionary$source_column == "group_code"] <-
  "study_group"
epi_eda_dictionary_validate(dictionary, catalogues)
```

PostgreSQL NULL is counted in exclusive row partitions and never returned as a mapping value. Literal empty text, `"NA"` and `"NULL"` remain ordinary observed values. Review this technical evidence alongside source documentation, then curate catalogue definitions or the EDA specification explicitly; the profiler changes neither source data nor dictionary semantics.

## 3. Generate review-only QC prompts

`epi_eda_qc_proposals()` accepts the reviewed specification plus a caller-created, persisted key map containing exactly `name` and `variable_key`. Each key is an opaque identifier beginning `var_` followed by 16 to 64 lower-case letters or digits. The package uses names only for exact input lookup and returns no name/key crosswalk, so reuse the same private map when comparing equivalent in-memory and PostgreSQL sources.

```{r}
variable_keys <- data.frame(
  name = spec$name,
  variable_key = c(
    "var_0000000000000001",
    "var_0000000000000002",
    "var_0000000000000003",
    "var_0000000000000004"
  ),
  stringsAsFactors = FALSE
)

source_before <- received_data
qc <- epi_eda_qc_proposals(received_data, spec, variable_keys)
stopifnot(identical(received_data, source_before))
qc
qc$evidence
qc$proposals
```

The `evidence` table has one `evidence_state = "descriptive"` row per specification row. It reports aggregate counts and, for numeric or integer variables, finite observed extrema and type-7 Tukey 1.5-IQR fences. These values describe only the profiled snapshot: they are not scientific validity limits and do not replace semantic `min` or `max`. Explicit `id` or `identifier` roles are not profiled, high-cardinality variables contribute a scalar distinct count without returned values, and incompatible or unsupported storage receives a fixed unavailable status.

The separate `proposals` table contains only `proposal_state = "pending"` prompts. An observed numeric or integer variable containing exactly both 0 and 1 may receive a binary candidate, finite observations beyond a Tukey fence may receive candidate screening bounds, and a blank unit may prompt review. The function never reclassifies a variable, infers a unit or missing-value code, enumerates general observed levels, changes data or the semantic dictionary, creates a cleaning rule, or writes an output. Executable cleaning rules use the separate contract below; editing `proposal_state` does not make this result executable.

The same call accepts an `epi_eda_postgres_source` and performs scalar aggregate queries in one read-only repeatable-read snapshot. QC proposals are not added automatically to intake, report or database bundles. Aggregate counts and extrema can still be sensitive in context, so review disclosure risk before saving or sharing either table.

## 4. Apply neutral cleaning rules

`epi_eda_cleaning_rules()` requires exactly six ordered technical fields: `variable_key`, `declared_type`, `valid_min`, `valid_max`, `allowed_values` and `missing_codes`. Executable bounds are named `valid_min` and `valid_max`; descriptive dictionary `min` and `max` remain metadata. The caller remains responsible for deciding whether each rule is scientifically justified.

```{r}
cleaning_rules <- epi_eda_cleaning_rules(data.frame(
  variable_key = variable_keys$variable_key[variable_keys$name == "age"],
  declared_type = "integer",
  valid_min = 0,
  valid_max = 120,
  allowed_values = "",
  missing_codes = "",
  stringsAsFactors = FALSE
))

cleaning <- epi_eda_apply_cleaning_rules(
  received_data,
  cleaning_rules,
  variable_keys
)
analysis_data <- cleaning$data
cleaning$audit$summary
cleaning$audit$variables
stopifnot(identical(received_data, source_before))
```

Numeric or integer values outside inclusive declared bounds become typed missing values. Categorical or binary values outside a populated semicolon-delimited `allowed_values` set become typed missing, and declared `missing_codes` become typed missing for any supported rule. Every rule and source mapping is validated before transformation. The result preserves data-frame row count, order, row names, column order and source storage and returns aggregate before/after and transition counts linked only by opaque variable key.

The released `epi_eda_approved_rules()` symbol remains temporarily as a deprecated redirect and will be removed in the next breaking release. It accepts only the same six-field schema, warns, and returns `epi_eda_cleaning_rules`; input using the former eight-field schema fails rather than having fields ignored or removed.

Supply both `output_path` and explicit `output_format = "csv"` or `"rds"` to publish the same complete processed data without replacing an existing file. For PostgreSQL, pass an `epi_eda_postgres_source` plus a new `destination_schema` and `destination_table`; server-side transformations, destination collision checks, dimension checks and missingness reconciliation run in one write transaction. Failure leaves no destination table. PostgreSQL tables have no physical row-order contract, but the operation neither filters nor explicitly reorders source rows.

```{r eval = FALSE}
file_result <- epi_eda_apply_cleaning_rules(
  received_data,
  cleaning_rules,
  variable_keys,
  output_path = "processed-data.rds",
  output_format = "rds"
)

database_result <- epi_eda_apply_cleaning_rules(
  source,
  cleaning_rules,
  variable_keys,
  destination_schema = "analysis",
  destination_table = "processed_observations"
)
```

The audit omits source names, paths, relation identities, allowed or disallowed values, bounds, missing codes and process metadata. Its aggregate counts and deterministic rule hash can still be sensitive in context, so review them before sharing. Text, date and datetime rules, RData, Parquet, overwriting, row filtering, imputation, winsorisation and civil-date derivation are outside this contract.

## 5. Derive explicitly reviewed civil dates

Use civil-date derivation only after external analyst review establishes that a local timestamp field means a civil calendar date. Midnight observations, a field name and a PostgreSQL type do not establish that meaning. `epi_eda_approved_civil_dates()` requires an exact schema with `operation_state = "approved"`, `declared_semantics = "civil_date"`, `preserve_source = TRUE`, `require_midnight = TRUE`, an opaque source key and an opaque caller-managed approval reference.

```{r}
local_records <- data.frame(
  local_stamp = c("2024-02-29 00:00:00", "2024-12-31 00:00:00.000", NA_character_),
  stringsAsFactors = FALSE
)
local_keys <- data.frame(
  name = "local_stamp",
  variable_key = "var_0000000000000001",
  stringsAsFactors = FALSE
)
approved_civil_dates <- epi_eda_approved_civil_dates(data.frame(
  source_variable_key = "var_0000000000000001",
  derived_name = "reviewed_date",
  operation_state = "approved",
  declared_semantics = "civil_date",
  preserve_source = TRUE,
  require_midnight = TRUE,
  approval_id = "approval_0000000000000001",
  stringsAsFactors = FALSE
))

local_before <- local_records
civil_dates <- epi_eda_derive_civil_dates(
  local_records,
  approved_civil_dates,
  local_keys
)
civil_dates$data
stopifnot(identical(local_records, local_before))
```

In memory, local timestamps must be strict timezone-free character values in `YYYY-MM-DD HH:MM:SS` form with an optional decimal fraction; `POSIXct`, `POSIXlt`, offsets and zones are rejected rather than interpreted. Fractional-zero midnight is valid, but any non-zero hour, minute, second or fractional digit blocks the complete operation. The error reports only the aggregate affected-value count. Missing timestamps produce missing `Date` values, and calendar boundaries are validated before derivation.

The PostgreSQL path accepts only `timestamp without time zone`, validates exact midnight with a scalar server-side aggregate and derives with a local `::date` cast in one write transaction. It does not collect timestamps, assign a timezone or relax the instant-oriented preparation rejection. The source column remains in the new table unchanged, while the new derived column uses PostgreSQL `date` storage.

```{r eval = FALSE}
database_dates <- epi_eda_derive_civil_dates(
  source,
  approved_civil_dates,
  local_keys,
  destination_schema = "analysis",
  destination_table = "reviewed_dates"
)
```

Supply both `output_path` and `output_format = "csv"` or `"rds"` for a new file. Existing files and PostgreSQL relations are never replaced, and failure leaves no partial destination. Returned audits omit source and derived names, paths, relation identities, timestamps and approval references; aggregate counts and hashes can still be sensitive in context.

## 6. Audit and apply data-frame preparation

Preparation is a data-frame stage. Audit mode returns the original data and factual findings. Apply mode is all-or-nothing: a blocking finding returns the original data, no `schema_after`, and `metadata$overall_status == "blocked"`; a successful apply returns prepared data and `metadata$overall_status == "prepared"`. `min` and `max` are descriptive metadata rather than recoding rules.

```{r}
assessment <- epi_eda_prepare(analysis_data, spec, mode = "audit")
assessment$metadata$overall_status
assessment$audit

prepared <- epi_eda_prepare(analysis_data, spec, mode = "apply")
stopifnot(prepared$metadata$overall_status == "prepared")
prepared$schema_after
```

`epi_eda_intake_run()` composes this audit with preparation, summaries, optional stratification and an owned bundle. Use `prepare = "audit"` to stop after a successful audit, `prepare = "apply"` to apply reviewed transformations, or `prepare = "none"` only when the audit requires no changes. Intake statuses are `audit_complete`, `complete` and `blocked`; invalid arguments and unsafe output collisions are errors.

```{r}
audit_dir <- tempfile("episcout-audit-intake-")
audit <- epi_eda_intake_run(
  received_data,
  spec,
  output_dir = audit_dir,
  prepare = "audit",
  render = FALSE
)
audit$status

intake_dir <- tempfile("episcout-complete-intake-")
intake <- epi_eda_intake_run(
  received_data,
  spec,
  output_dir = intake_dir,
  prepare = "apply",
  strata = "study_group",
  render = FALSE
)
intake$status
intake$stratified$groups
intake$table1
```

With `spec = NULL`, intake generates, saves and uses a storage-derived scaffold as `specification.csv`; use that shortcut only when its undeclared semantic fields are acceptable for the current run. Intake creates a missing output directory. Its five-column manifest has `artifact`, `type`, `path`, `status` and `checksum_md5`, and manifest paths are relative to the exact output root. It never writes source or prepared rows, row previews, raw free-text examples, identifier values or pseudonymisation bridge tables.

PostgreSQL-backed EDA has no `epi_eda_prepare()` audit/apply mode. Use `epi_eda_apply_cleaning_rules()` only for the declared bounded operations above, then construct a new PostgreSQL source from the processed destination. Resolve every other storage compatibility or semantic recoding need in a reviewed relation or view before EDA.

## 7. Run summaries, plots and optional stratification

The canonical summary contains `variables`, `numeric`, `categorical`, `text`, `temporal` and `skipped`. Every specification row remains accounted for. Numeric summaries separate finite values from infinities, categorical summaries retain declared zero-count levels, text summaries contain length and cardinality aggregates rather than observed strings, and temporal summaries state their range units.

```{r}
results <- epi_eda_run(prepared$data, spec)
names(results$summaries)
results$summaries$variables
results$categorical_display

stratified <- epi_eda_profile_stratified(
  prepared$data,
  spec,
  strata = "study_group",
  max_levels = 10
)
table1 <- epi_eda_table1(stratified)
table1
```

`epi_eda_profile_stratified()` accepts either a data frame or an `epi_eda_postgres_source` for exactly one declared categorical or binary stratifier. Its omitted `max_levels = Inf` default preserves the released unbounded result. A supplied finite positive whole-number bound applies independently to the ordinary stratum domain, every compared categorical or binary domain and each declared-plus-observed union. One value over the bound fails the complete call without truncation; PostgreSQL collection is limited in the query path. PostgreSQL stratification runs in one read-only repeatable-read transaction and returns only grouped or scalar aggregates. It does not collect the analysis-value vector required by Shapiro-Wilk, so PostgreSQL stratified numeric `shapiro_p` values are unavailable and this limitation is recorded in result metadata. The returned object works unchanged with `epi_eda_table1()` and `epi_eda_categorical_display()`.

Explicit numeric associations remain caller-selected. `epi_eda_profile_spearman()` takes an ordered two-column character data frame, rejects identifiers and unsupported or duplicate pairs, and reranks each eligible pair independently after excluding standard missing values, declared missing codes and non-finite values. PostgreSQL returns one aggregate row per pair from one read-only repeatable-read transaction; neither backend returns a p-value, confidence interval, threshold or interpretation.

```{r}
numeric_pairs <- data.frame(
  left = "age",
  right = "longitude",
  stringsAsFactors = FALSE
)
spearman <- epi_eda_profile_spearman(prepared$data, spec, numeric_pairs)
spearman
```

`epi_eda_cramers_v()` consumes long-form aggregate contingency counts rather than source rows. The required columns are `left`, `right`, `row_level`, `column_level` and `n`. Explicit zero-count rows retain declared domain evidence, while zero-total margins are removed from the active dimensions before the coefficient is calculated. Missing row/column combinations among supplied levels are treated as zero cells. Zero totals and one-dimensional active tables return typed unavailable evidence.

```{r}
contingency_counts <- data.frame(
  left = "group",
  right = "outcome",
  row_level = rep(c("r1", "r2"), each = 2L),
  column_level = rep(c("c1", "c2"), 2L),
  n = c(30, 10, 10, 50),
  stringsAsFactors = FALSE
)
epi_eda_cramers_v(contingency_counts)
```

To reuse a bounded stratified categorical component, retain ordinary non-missing group and level rows, map `group_value`, `name`, `level` and `n` to `row_level`, `right`, `column_level` and `n`, and repeat `stratified$metadata$strata` as `left`. This transformation uses the existing aggregate object and requires no source-row access or second PostgreSQL query. Pair selection, labels, ranking, presentation and scientific interpretation remain caller responsibilities.

`epi_eda_categorical_display()` is the shared aggregate calculation behind categorical presentation. Every row retains an integer `numerator`, integer `denominator`, numeric `proportion`, percentage-basis fields and explicit missingness treatment. Standard missing values and declared missing codes are combined into one missing-level row. A zero denominator remains integer zero and its proportion is `NA_real_`; it is never presented as zero percent.

```{r}
compatibility <- epi_eda_categorical_display(stratified)
column <- epi_eda_categorical_display(stratified, basis = "column")
row <- epi_eda_categorical_display(stratified, basis = "row")
overall <- epi_eda_categorical_display(stratified, basis = "overall")

epi_eda_table1(stratified, basis = "column")
```

The default `compatibility` basis uses observed non-missing values for ordinary levels and all rows in the corresponding group for the missing level. `column` uses all rows in each group, `row` distributes each level across non-Overall strata, and `overall` uses the included analysis population. Row percentages require a stratified result. When outputs are requested, `epi_eda_run()` and intake write the compatibility calculation as `categorical_display.csv`; Table 1 records the numeric denominator behind each displayed count and percentage.

## 8. Request declared point maps

Mapping is off by default. `maps = TRUE` requests one geometry-only point map per declared pair; each unique `map_vars` entry requests one additional thematic map per pair. Themes may be numeric, integer, categorical, binary or text. Date and datetime themes are unsupported, and no thematic variable is selected automatically.

```{r, fig.width = 6, fig.height = 4}
mapped <- epi_eda_run(
  prepared$data,
  spec,
  maps = TRUE,
  map_vars = "study_group",
  max_map_points = 10000L
)
mapped$geo
mapped$map_inventory
mapped$maps[["map-p001-geometry"]]
mapped$maps[["map-p001-v002"]]
```

Maps are created only when every row in a pair is complete, finite, within EPSG:4326 ranges when applicable, and the source has at most `max_map_points` rows. A failed pair records `not_created`, preserves the rest of EDA, and is never partially mapped or truncated. Map readiness is structural QA, not evidence that coordinates represent the intended place, person, time or unit, and not disclosure approval.

The same map controls are available on intake, the data-frame report renderer and the PostgreSQL database run. Bundle-producing runs write deterministic SVGs under `maps/`; reports embed created maps and their inventory. Requested point maps contain individual locations and may contain thematic values, so an output with maps is not aggregate-only. Integrated EDA creates coordinate-derived points only; use the [geospatial mapping guide](geospatial-mapping-primer.html) for existing `sf`, line, polygon, boundary, basemap and PostGIS geometry workflows.

## 9. Run PostgreSQL EDA in one snapshot

PostgreSQL 17 or later is required. The caller supplies an open, idle RPostgres connection and separate schema and relation identifiers. Ordinary tables, partitioned tables, views, materialised views and foreign tables are supported. Arbitrary SQL, dotted relation names, temporary objects, generic DBI connections and caller-managed transactions are not accepted.

```{r eval = FALSE}
con <- DBI::dbConnect(
  RPostgres::Postgres(),
  host = Sys.getenv("PGHOST"),
  port = as.integer(Sys.getenv("PGPORT", unset = "5432")),
  dbname = Sys.getenv("PGDATABASE"),
  user = Sys.getenv("PGUSER")
)
source <- epi_eda_postgres_source(con, "study", "observations")

db_results <- epi_eda_db_run(
  source,
  spec,
  output_dir = "YYYYMMDD_eda_cycle",
  plots = TRUE,
  maps = FALSE,
  strata = "study_group",
  table1_basis = "column",
  layout = "delivery"
)
DBI::dbDisconnect(con)
```

The six direct profilers `epi_eda_check_schema()`, `epi_eda_profile_missing()`, `epi_eda_profile_summaries()`, `epi_eda_profile_plots()`, `epi_eda_profile_stratified()` and `epi_eda_profile_spearman()` also accept a PostgreSQL source; each direct call owns its own read-only repeatable-read transaction. `epi_eda_db_run()` performs catalogue revalidation, schema and missingness checks, canonical summaries, identifier QA, compact plot preparation, optional stratification and map-readiness QA in one read-only repeatable-read snapshot. Rendering and atomic filesystem publication occur after the snapshot ends. Supplying `strata` adds manifest-owned stratified components and Table 1 while leaving the default bundle unchanged. Stratified database bundles reject source-row maps and disable Shapiro-Wilk in both Overall and grouped numeric summaries to preserve the no-analysis-vector boundary.

Without maps, the database bundle contains aggregates, the caller-authored specification and aggregate plot inputs only. When maps are requested, coordinates and selected thematic observations are collected only for map-ready pairs within the inclusive row limit. The query projects only the required coordinates and themes, applies a defensive `max_map_points + 1` limit, and fails rather than truncating or silently dropping rows. The bundle never contains a source-row table, arbitrary SQL, query parameters, credentials or connection attributes; database and driver logs remain the caller's infrastructure responsibility.

Use `plot_style` only for creation-time presentation changes such as an approved theme. The callback receives a completed `ggplot` and compact metadata (`name`, `label`, declared `type`, `plot_type` and aggregate counts), and must return one `ggplot`; it receives no rows, connection, SQL or credentials. For `epi_eda_db_run(plots = TRUE)`, pair it with a stable, non-secret `plot_style_id`. The identifier is recorded in run metadata and must match before an owned bundle is overwritten. Styling runs after the repeatable-read snapshot closes and before staged SVG publication, so the returned and written plots use the same style. `plots = FALSE` does not call the callback.

## 10. Deliver the owned output root

For data-frame intake, `render = TRUE` creates `report.html` from the saved bundle components. The exact output root is workflow-owned: use `overwrite = TRUE` only to replace a prior valid owned intake bundle. Do not move individual files out of the root without preserving their manifest relationship.

For a lower-level data-frame report, create an empty output directory and pass already prepared data to `epi_eda_render_report()`. It writes machine-readable CSV components and `eda-report.html`, but it does not create a checksum manifest and does not run preparation.

```{r eval = FALSE}
output_dir <- file.path(tempdir(), "episcout-eda")
dir.create(output_dir)
epi_eda_render_report(
  prepared$data,
  spec,
  output_dir,
  maps = TRUE,
  map_vars = "study_group"
)
```

For PostgreSQL delivery, `layout = "delivery"` creates `README.md` and `reports/eda-report.html` as human entry points. `QA_QC/` contains aggregate checks and summaries, `plot_data/` contains compact aggregate plot inputs but no map coordinates, themes or source rows, `plots/` and `maps/` contain requested deterministic SVGs, and `run_manifests/` contains source metadata, timings, delivery metadata and the five-column checksum manifest. The caller supplies the exact root; a `YYYYMMDD_eda_cycle` name is recommended but not inferred. The default `layout = "bundle"` retains the flat aggregate bundle and has no automatic HTML entry point.

An existing flat or delivery database bundle can be rendered without a live connection. The renderer re-reads and validates every on-disk manifest-owned artefact; a run result contributes only its `output_dir`. Use `overwrite = TRUE` only to replace an existing owned README and report.

```{r eval = FALSE}
report <- epi_eda_render_db_report("postgres-eda-bundle")
```

Retain and transfer the complete owned root. The report is a view of the validated canonical CSV, SVG, checksum and provenance evidence; it does not replace those artefacts. A complete run has `status == "complete"`. A database-run validation or publication failure raises an error rather than returning an intake-style blocked bundle.

Neither a report nor an aggregate-only bundle is automatically safe to share. Review caller-authored dictionary metadata, categories, small cells, plots and any requested point maps. Pseudonymised observations remain restricted personal data: pseudonymisation replaces or separates direct identifiers under a controlled linkage process, but it does not make the observations anonymous or automatically disclosure-controlled.

## 11. Start from the file-based scaffold

Create a starter project, edit `metadata/data_dictionary.csv`, place an authorised input at `data/input.csv`, and follow the copied README. The scaffold offers direct and optional `targets` paths to the same lower-level data-frame renderer; it does not manage PostgreSQL connections or preparation decisions.

```{r eval = FALSE}
epi_eda_create_project("my-eda-project")
```

Synthetic generation in the scaffold is for pipeline preparation and testing only. It is not suitable for inference, anonymity or disclosure control. The lower-level `epi_eda_*` functions remain available for ordinary scripts when an owned intake bundle or project scaffold is unnecessary.
