---
title: "Specification-first exploratory data analysis"
author: "Antonio Berlanga-Taylor"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Specification-first exploratory data analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_ggplot2
)
```

```{r eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("This vignette needs ggplot2 to run the plotting step. Install ggplot2 to run the examples.")
```

Specification-first EDA starts with a data dictionary rather than with ad hoc
analysis code. The same specification can be used while data access is pending,
then reused with real data once they arrive.

The MVP workflow is designed to:

- validate expected variables;
- generate simple synthetic data for pipeline preparation;
- profile missingness;
- produce descriptive summaries;
- create one basic plot per variable;
- optionally write machine-readable outputs and render an HTML report.

Synthetic data generated by `episcout` are for workflow preparation and testing
only. They are not suitable for inference, publication, privacy protection or
external data release.

## Create a specification

The specification is usually stored as `metadata/data_dictionary.csv`. It must
include `name`, `label`, `type` and `role`.

```{r}
library(episcout)

spec_data <- data.frame(
  name = c("age", "sex", "death", "follow_up_days"),
  label = c(
    "Age at baseline",
    "Sex at birth",
    "Death during follow-up",
    "Follow-up time"
  ),
  type = c("numeric", "categorical", "binary", "integer"),
  role = c("covariate", "covariate", "outcome", "outcome"),
  units = c("years", "", "", "days"),
  levels = c("", "Female;Male;Unknown", "0;1", ""),
  min = c(18, NA, 0, 0),
  max = c(110, NA, 1, 3650),
  missing_codes = c("", "", "", ""),
  required = c(TRUE, TRUE, TRUE, TRUE),
  group = c("demographics", "demographics", "outcomes", "outcomes"),
  description = c(
    "Age in years at study baseline.",
    "Recorded sex at birth.",
    "Binary outcome indicator.",
    "Observed follow-up time in days."
  ),
  stringsAsFactors = FALSE
)

spec <- epi_eda_spec(spec_data)
spec
```

## Prepare the workflow before data access

Use the specification to generate deterministic synthetic data, then run the
same workflow that will later be used for real data.

```{r}
synthetic_results <- epi_eda_run(
  data = NULL,
  spec = spec,
  synthetic = TRUE,
  n = 100,
  seed = 1
)

names(synthetic_results)
synthetic_results$metadata
synthetic_results$schema
head(synthetic_results$missing)
```

The `summaries` component contains separate machine-readable tables for numeric
and categorical variables.

```{r}
synthetic_results$summaries$numeric
synthetic_results$summaries$categorical
```

Plots are returned as ggplot objects without being printed by the workflow.

```{r, fig.width = 5, fig.height = 3}
synthetic_results$plots$age
```

## Replace synthetic data with real data

When real data are available, keep the same specification and change the data
source. This example creates a small in-memory data frame to show the shape of
the call.

```{r}
real_data <- data.frame(
  age = c(42, 57, 61, NA),
  sex = c("Female", "Male", "Female", "Unknown"),
  death = c("0", "1", "0", "0"),
  follow_up_days = c(365L, 420L, 800L, 90L),
  stringsAsFactors = FALSE
)

real_results <- epi_eda_run(
  data = real_data,
  spec = spec
)

real_results$metadata
real_results$missing
```

## Write outputs and render a report

`epi_eda_run()` writes CSV outputs when `output_dir` is supplied. The directory
must already exist.

```{r}
output_dir <- file.path(tempdir(), "episcout-eda")
dir.create(output_dir, showWarnings = FALSE)

epi_eda_run(
  data = real_data,
  spec = spec,
  output_dir = output_dir
)

list.files(output_dir)
```

If `rmarkdown` is installed, `epi_eda_render_report()` runs the workflow and
renders a bundled HTML report from the workflow outputs.

```{r eval = FALSE}
epi_eda_render_report(
  data = real_data,
  spec = spec,
  output_dir = output_dir
)
```

## Start a project scaffold

For a file-based project, create a starter scaffold and then edit the copied
data dictionary.

```{r eval = FALSE}
epi_eda_create_project("my-eda-project")
```

The scaffold includes:

- `metadata/data_dictionary.csv`;
- `data/` for real input data;
- `outputs/` for generated workflow files;
- `_targets.R` for optional targets-based execution;
- `reports/eda.qmd` for direct report rendering.

The reporting and scaffold workflow is optional. You can use the core
`epi_eda_*` functions directly in ordinary R scripts.
