---
title: "Specification-first exploratory data analysis"
author: "Antonio Berlanga-Taylor"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Specification-first exploratory data analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_ggplot2
)
```

```{r eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("This vignette needs ggplot2 to run the plotting step. Install ggplot2 to run the examples.")
```

Specification-first EDA starts with a data dictionary rather than with ad hoc analysis code. The same specification can be used while data access is pending, then reused with real data once they arrive.

For an executable end-to-end example that also covers duplicate review, longitudinal PostgreSQL pseudonymisation and the handoff into this workflow, open the installed `examples/db-to-report/walkthrough.R` script and run it section by section.

The specification-first workflow is designed to:

- validate expected variables;
- generate simple synthetic data for pipeline preparation;
- profile missingness;
- produce descriptive summaries;
- create one basic plot per variable;
- optionally write machine-readable outputs and render an HTML report.

Synthetic data generated by `episcout` are for workflow preparation and testing only. They are not suitable for inference, publication, privacy protection or external data release.

## Start when data arrive before a dictionary

When an existing data frame arrives without a dictionary, `epi_eda_spec_scaffold()` creates a conservative draft from storage classes and aggregate structure. It does not enumerate observed values or infer scientific roles, units, missing sentinels, validation limits, requiredness or privacy status.

```{r}
library(episcout)

received_data <- data.frame(
  age = c(42, 57, 61, NA),
  study_group = c("A", "B", "A", "B"),
  stringsAsFactors = FALSE
)

draft_spec <- epi_eda_spec_scaffold(received_data)
draft_spec[, c(
  "name", "type", "candidate_type", "n_missing", "n_unique",
  "review_status"
)]

draft_path <- tempfile(fileext = ".csv")
write.csv(draft_spec, draft_path, row.names = FALSE, na = "")
```

Every draft row requires human review. In this neutral example, the review confirms roles and requiredness and promotes the character grouping candidate to a categorical declaration.

```{r}
reviewed_spec <- epi_eda_spec(draft_path)
reviewed_spec$role <- c("covariate", "exposure")
reviewed_spec$required <- c(TRUE, TRUE)
reviewed_spec$type[reviewed_spec$name == "study_group"] <- "categorical"
reviewed_spec$levels[reviewed_spec$name == "study_group"] <- "A"
reviewed_spec$description <- c(
  "Age recorded in the received dataset.",
  "Reviewed study grouping."
)
reviewed_spec$review_status <- "reviewed"
reviewed_spec <- epi_eda_spec(reviewed_spec)

assessment <- epi_eda_prepare(received_data, reviewed_spec, mode = "audit")
assessment$audit[assessment$audit$status == "blocking", ]
```

The audit counts an unexpected grouping observation without including its value. Review of the source and dictionary confirms that both groups are valid, so the declaration—not the observation—is corrected before apply.

```{r}
reviewed_spec$levels[reviewed_spec$name == "study_group"] <- "A;B"
reviewed_spec <- epi_eda_spec(reviewed_spec)

prepared <- epi_eda_prepare(received_data, reviewed_spec, mode = "apply")
prepared$metadata
prepared$schema_before
prepared$schema_after

received_results <- epi_eda_run(prepared$data, reviewed_spec)
names(received_results$summaries)
received_results$summaries$variables
```

Audit is assessment; apply is a separate all-or-nothing mutation of a returned copy. Missing sentinels are reviewed missingness declarations, whereas unexpected values require source or level review and are never silently converted to missing. Observed `min` and `max` evidence is not a validation range and preparation does not clamp values. Factor levels are storage metadata and can themselves contain sensitive codes, so inspect the draft before saving or sharing it. Factor metadata that cannot round-trip through the package's semicolon-delimited format is refused rather than silently changed. The same v1 encoding cannot declare empty, whitespace-only or semicolon-containing missing sentinels. Database inventory workflows use `epi_eda_dictionary_scaffold()` instead.

Preparation neither identifies personal information nor anonymises data. `epi_sec_pseudonym()` creates a controlled single-vector bridge table but does not maintain stable tokens across later calls or rewrite related tables. For restricted PostgreSQL sources, use the dedicated `longitudinal-pseudonymisation` vignette to review a value-free linkage contract, audit before writing and hand the resulting output dictionary back into EDA. Pseudonymised data remain restricted personal data; they are not anonymous or automatically disclosure-controlled.

## Use the stage-gated new-dataset workflow

`epi_eda_intake_run()` connects the review, audit, preparation and descriptive stages without treating a generated scaffold as approval. Its first call writes a scaffold and review guide, then returns before schema or analytical work.

```{r}
first_dir <- tempfile("episcout-first-intake-")
first_run <- epi_eda_intake_run(
  received_data,
  output_dir = first_dir,
  render = FALSE
)

first_run$status
first_run$stage
first_run$manifest[first_run$manifest$status == "created", c("path", "sensitivity")]
```

The analyst edits the scaffold outside the run. This example deliberately declares only one observed group so the second call demonstrates a value-free blocker; the message identifies the affected variable and policy but does not copy the unexpected value.

```{r}
intake_spec <- epi_eda_spec(file.path(first_dir, "spec_scaffold.csv"))
intake_spec$role <- c("covariate", "exposure")
intake_spec$required <- TRUE
intake_spec$type[intake_spec$name == "study_group"] <- "categorical"
intake_spec$levels[intake_spec$name == "study_group"] <- "A"
intake_spec$review_status <- "reviewed"

blocked_dir <- tempfile("episcout-blocked-audit-")
blocked_run <- epi_eda_intake_run(
  received_data,
  intake_spec,
  output_dir = blocked_dir,
  prepare = "audit",
  render = FALSE
)

blocked_run$status
blocked_run$messages[, c("stage", "severity", "subject", "recommended_action")]
```

After confirming the second group with the source documentation, the analyst corrects the specification and repeats the audit. Audit mode never claims to have transformed data.

```{r}
intake_spec$levels[intake_spec$name == "study_group"] <- "A;B"

audit_dir <- tempfile("episcout-complete-audit-")
audit_run <- epi_eda_intake_run(
  received_data,
  intake_spec,
  output_dir = audit_dir,
  prepare = "audit",
  render = FALSE
)

audit_run$status
audit_run$preparation_audit
```

The final call explicitly applies the approved character-to-factor preparation in memory, calculates the canonical six-component profile, calculates the optional grouped profile and Table 1, and renders `report.html` from the saved CSV components. It never writes the prepared rows.

```{r}
final_dir <- tempfile("episcout-complete-intake-")
final_run <- epi_eda_intake_run(
  received_data,
  intake_spec,
  output_dir = final_dir,
  prepare = "apply",
  strata = "study_group"
)

final_run$status
final_run$stage
names(final_run$summary)
final_run$stratified$groups
final_run$table1
list.files(final_dir)
```

Statuses are `review_required`, `blocked`, `audit_complete` or `complete`; `stage` records the last completed data-processing stage even when an optional report fails. Reuse an existing non-empty run directory only with `overwrite = TRUE`; overwrite is accepted only for a prior valid intake bundle with no unowned entries, and only registered workflow files are replaced. Relative manifest/report links remain valid if the whole bundle moves.

Explicit roles `id` and `identifier` retain aggregate missingness but are policy-skipped from typed profiles and value-bearing plots, so observed identifier values do not enter analytical components, CSV summaries or the report. This is not automatic PII detection. Caller-supplied specification fields, variable names, category/group labels and small aggregate cells may remain sensitive and are retained as review-required metadata, so every bundle requires disclosure review. No bridge table, row-level input or prepared data is written, and the report is not evidence of de-identification or publication safety.

## Create a specification

The specification is usually stored as `metadata/data_dictionary.csv`. It must include `name`, `label`, `type` and `role`.

```{r}
spec_data <- data.frame(
  name = c("age", "sex", "death", "follow_up_days"),
  label = c(
    "Age at baseline",
    "Sex at birth",
    "Death during follow-up",
    "Follow-up time"
  ),
  type = c("numeric", "categorical", "binary", "integer"),
  role = c("covariate", "covariate", "outcome", "outcome"),
  units = c("years", "", "", "days"),
  levels = c("", "Female;Male;Unknown", "0;1", ""),
  min = c(18, NA, 0, 0),
  max = c(110, NA, 1, 3650),
  missing_codes = c("", "Unknown", "", ""),
  required = c(TRUE, TRUE, TRUE, TRUE),
  group = c("demographics", "demographics", "outcomes", "outcomes"),
  description = c(
    "Age in years at study baseline.",
    "Recorded sex at birth.",
    "Binary outcome indicator.",
    "Observed follow-up time in days."
  ),
  stringsAsFactors = FALSE
)

spec <- epi_eda_spec(spec_data)
spec
```

The optional `missing_codes` column accepts semicolon-separated sentinel values. These values are counted as missing in the missingness profile and excluded from observed summaries and plots. Categorical summaries report `p` over all rows and `p_observed` over observed non-missing rows.

## Prepare the workflow before data access

Use the specification to generate deterministic synthetic data, then run the same workflow that will later be used for real data.

```{r}
synthetic_results <- epi_eda_run(
  data = NULL,
  spec = spec,
  synthetic = TRUE,
  n = 100,
  seed = 1
)

names(synthetic_results)
synthetic_results$metadata
synthetic_results$schema
head(synthetic_results$missing)
```

Schema output reports variable presence in `status` and separately reports whether observed storage is `compatible`, `coercible`, `incompatible` or `not_applicable` in `type_status`, with an explanation in `type_reason`. This check is descriptive: it does not coerce data or stop the workflow for an incompatible type.

The `summaries` component uses one canonical contract with `variables`, `numeric`, `categorical`, `text`, `temporal` and `skipped` tables. The variables table accounts for every specification row, including missing or incompatible variables.

```{r}
names(synthetic_results$summaries)
synthetic_results$summaries$variables
synthetic_results$summaries$numeric
synthetic_results$summaries$categorical
```

Numeric summaries exclude infinities from finite analytical values while recording them in variable coverage. Categorical summaries retain declared zero-count levels, append unexpected observed values and expose both total-row and observed-value denominators. Temporal summaries use days for dates and seconds for datetimes.

```{r}
synthetic_results$summaries$text
synthetic_results$summaries$temporal
synthetic_results$summaries$skipped
```

Plots are returned as ggplot objects without being printed by the workflow. Numeric, text and temporal plots are built from compact 30-bin tables; text uses Unicode character lengths and never raw strings. Explicit identifier roles have named `NULL` entries.

```{r, fig.width = 5, fig.height = 3}
synthetic_results$plots$age
```

## Run aggregate-only EDA in PostgreSQL

PostgreSQL 17 or later is the supported server-backed EDA source. The caller creates and owns an RPostgres connection, then identifies one ordinary table, partitioned table, view, materialized view or foreign table with separate schema and relation names. Arbitrary SQL, dotted names, temporary objects, generic DBI connections and connections already inside a transaction are rejected.

```{r eval = FALSE}
con <- DBI::dbConnect(
  RPostgres::Postgres(),
  host = Sys.getenv("PGHOST"),
  dbname = Sys.getenv("PGDATABASE"),
  user = Sys.getenv("PGUSER")
)

source <- epi_eda_postgres_source(
  con,
  schema = "eda_fixture",
  relation = "observations"
)

epi_eda_check_schema(source, spec)
epi_eda_profile_missing(source, spec)
epi_eda_profile_summaries(source, spec)
epi_eda_profile_plots(source, spec)

db_results <- epi_eda_db_run(
  source,
  spec,
  output_dir = "postgres-eda-bundle",
  plots = TRUE,
  max_plot_levels = 20L
)

DBI::dbDisconnect(con)
```

Each direct profiler owns one read-only repeatable-read transaction. The complete database run uses one stable snapshot for catalogue revalidation, schema, missingness, canonical summaries, aggregate identifier QA and compact plot data. It ends the transaction before SVG rendering and staged filesystem publication. `timestamp without time zone` is incompatible because storage does not establish an instant or DST policy; expose a caller-reviewed `timestamp with time zone` cast in a view instead of relying on the session timezone.

The manifest-owned bundle contains the reviewed specification, relation identity, complete categorical frequencies, small aggregates, identifier QA, plot inventory, query timings without values or SQL, and optional SVGs. It contains no full rows, previews, raw text, observed identifier values, credentials, connection attributes or query parameters. This boundary is not de-identification: declared levels and missing sentinels in the caller-authored specification, names, categories, small cells and plots may be sensitive. Review the complete bundle before sharing, and apply PostgreSQL, driver, administrator, logging, backup and disclosure controls outside episcout.

## Replace synthetic data with real data

When real data are available, keep the same specification and change the data source. This example creates a small in-memory data frame to show the shape of the call.

```{r}
real_data <- data.frame(
  age = c(42, 57, 61, NA),
  sex = c("Female", "Male", "Female", "Unknown"),
  death = c("0", "1", "0", "0"),
  follow_up_days = c(365L, 420L, 800L, 90L),
  stringsAsFactors = FALSE
)

real_preparation <- epi_eda_prepare(real_data, spec, mode = "apply")
stopifnot(real_preparation$metadata$overall_status == "prepared")

real_results <- epi_eda_run(
  data = real_preparation$data,
  spec = spec
)

real_results$metadata
real_results$missing
```

## Describe the cohort by a reviewed group

Calculation remains separate from presentation. The stratified result uses the same canonical summary cores for Overall and every group, while retaining declared zero-count groups and levels.

```{r}
stratified <- epi_eda_profile_stratified(
  real_preparation$data,
  spec,
  strata = "sex"
)

stratified$groups
stratified$numeric
stratified$categorical

table1 <- epi_eda_table1(stratified)
table1
```

Ordinary categorical percentages use observed non-missing values; explicit missing rows use all rows in the group. `groups` and `metadata` identify missing and unexpected strata and account for omitted missing-stratum rows when `include_missing_stratum = FALSE`. Overall is calculated directly from the included population rather than from rounded group displays. Quantiles, finite-value rules and temporal units are inherited from the canonical EDA contract.

The Table 1 is a long-form plain data frame with no p-values. It does not suppress small cells and is not disclosure-controlled. Review categorical metadata and disclosure risk before sharing; text output contains aggregate length/cardinality diagnostics rather than observed examples.

## Write outputs and render a report

`epi_eda_run()` writes CSV outputs when `output_dir` is supplied. The directory must already exist.

```{r}
output_dir <- file.path(tempdir(), "episcout-eda")
dir.create(output_dir, showWarnings = FALSE)

epi_eda_run(
  data = real_data,
  spec = spec,
  output_dir = output_dir
)

list.files(output_dir)
```

If `rmarkdown` is installed, `epi_eda_render_report()` runs the workflow and renders a bundled HTML report from the workflow outputs.

```{r eval = FALSE}
epi_eda_render_report(
  data = real_data,
  spec = spec,
  output_dir = output_dir
)
```

## Start a project scaffold

For a file-based project, create a starter scaffold and then edit the copied data dictionary.

```{r eval = FALSE}
epi_eda_create_project("my-eda-project")
```

The scaffold includes:

- `metadata/data_dictionary.csv`;
- `data/` for real input data;
- `outputs/` for generated workflow files;
- `_targets.R` for optional targets-based execution;
- `reports/eda.qmd` for direct report rendering.

The reporting and scaffold workflow is optional. You can use the core `epi_eda_*` functions directly in ordinary R scripts.
