---
title: "PostgreSQL EDA Report"
output: html_document
params:
  bundle: null
---

```{r setup, include=FALSE}
bundle <- params$bundle
tables <- bundle$tables
show_table <- function(value, caption) {
  if (nrow(value) == 0L) {
    cat("No rows were produced for this component.\n\n")
  } else {
    print(knitr::kable(value, caption = caption, escape = TRUE))
    cat("\n\n")
  }
}
show_gallery <- function(value, label) {
  if (nrow(value) == 0L) {
    cat("No ", tolower(label), " SVGs were created.\n\n", sep = "")
    return(invisible(NULL))
  }
  for (index in seq_len(nrow(value))) {
    path <- paste0("../", value$path[[index]])
    cat("### ", label, " ", index, "\n\n", sep = "")
    cat("<object data=\"", path,
        "\" type=\"image/svg+xml\" style=\"max-width:100%;width:100%;height:32rem\">",
        "<a href=\"", path, "\">Open SVG</a></object>\n\n", sep = "")
  }
}
show_artifacts <- function(value) {
  created <- value[value$status == "created", , drop = FALSE]
  cat("<ul>\n")
  for (index in seq_len(nrow(created))) {
    path <- paste0("../", created$path[[index]])
    cat("<li><a href=\"", path, "\">", created$path[[index]],
        "</a> (", created$type[[index]], ")</li>\n", sep = "")
  }
  cat("</ul>\n")
}
```

# EDA delivery

This report is a human-facing view of the validated aggregate bundle. CSV,
SVG, checksums, and provenance remain the canonical evidence.

episcout creates the outputs explicitly requested by the analyst and does not
decide whether they may be shared.

## Run metadata

```{r metadata, echo=FALSE, results='asis'}
show_table(tables$run_metadata, "Run metadata")
```

## Source metadata

```{r source, echo=FALSE, results='asis'}
show_table(tables$source_metadata, "Source metadata")
```

## Specification and schema

```{r specification, echo=FALSE, results='asis'}
show_table(tables$specification, "Semantic specification")
show_table(tables$schema, "Schema checks")
```

## Missingness

```{r missing, echo=FALSE, results='asis'}
show_table(tables$missing, "Missing-value profile")
```

## Canonical summaries

```{r summaries, echo=FALSE, results='asis'}
components <- c("variables", "numeric", "categorical", "text", "temporal", "skipped")
for (component in components) {
  cat("### ", tools::toTitleCase(component), "\n\n", sep = "")
  show_table(tables[[paste0("summary_", component)]], paste(tools::toTitleCase(component), "summary"))
}
```

## Identifier QA

```{r identifier, echo=FALSE, results='asis'}
show_table(tables$identifier_qa, "Identifier QA")
```

## Declared coordinate-pair QA

```{r geo, echo=FALSE, results='asis'}
show_table(tables$geo_qa, "Coordinate-pair QA")
```

## Messages

```{r messages, echo=FALSE, results='asis'}
show_table(tables$messages, "Workflow messages")
```

## Categorical percentage companions

Frequency companions use the compatibility basis: ordinary levels are shown over observed values, while standard and declared-code missing values form one separate level over all rows. The plots remain count plots.

```{r categorical-companions, echo=FALSE, results='asis'}
if (length(bundle$frequency_companions) == 0L) {
  cat("No manifest-owned categorical frequency companions were produced.\n\n")
} else {
  for (label in names(bundle$frequency_companions)) {
    cat("### ", label, "\n\n", sep = "")
    show_table(
      bundle$frequency_companions[[label]],
      paste(label, "categorical numerator, denominator and proportion")
    )
  }
}
```

## Plot inventory

```{r plot-inventory, echo=FALSE, results='asis'}
show_table(tables$plot_inventory, "Plot inventory")
```

```{r plot-gallery, echo=FALSE, results='asis'}
show_gallery(bundle$plots, "Plot")
```

## Map inventory

```{r map-inventory, echo=FALSE, results='asis'}
show_table(tables$map_inventory, "Map inventory")
```

```{r map-gallery, echo=FALSE, results='asis'}
show_gallery(bundle$maps, "Map")
```

## Query timings

```{r timings, echo=FALSE, results='asis'}
show_table(tables$query_timings, "Query timings")
```

## Manifest

```{r artifact-index, echo=FALSE, results='asis'}
show_artifacts(bundle$manifest)
```

```{r manifest, echo=FALSE, results='asis'}
show_table(bundle$manifest, "Owned artifacts and checksums")
```
