---
title: "Specification-first EDA Report"
format: html
params:
  results: null
  synthetic: false
---

```{r setup, include=FALSE}
results <- params$results
mode_label <- if (isTRUE(params$synthetic)) "Synthetic" else "Real"
```

# EDA Report

Mode: `r mode_label`

episcout creates the outputs explicitly requested by the analyst and does not decide whether they may be shared.

```{r metadata, echo=FALSE}
knitr::kable(results$metadata, caption = "Workflow metadata")
```

## Schema

```{r schema, echo=FALSE}
knitr::kable(results$schema, caption = "Schema checks")
```

## Missingness

```{r missingness, echo=FALSE}
knitr::kable(results$missing, caption = "Missing-value profile")
```

## Summaries

The canonical summary accounts for every specification variable. Numeric statistics use finite observed values; `n_infinite` in the variables table records observed infinities excluded from those calculations. Categorical `p_total` values use all source rows, while `p_observed` values use non-missing observations. Temporal ranges state their units in `range_unit`. Variables that cannot be summarised remain visible with an actionable reason.

```{r summaries, echo=FALSE, results='asis'}
component_names <- c("variables", "numeric", "categorical", "text", "temporal", "skipped")
for (name in component_names) {
  summary_table <- results$summaries[[name]]
  heading <- paste(tools::toTitleCase(name), "Summaries")
  cat("### ", heading, "\n\n", sep = "")
  if (nrow(summary_table) == 0L) {
    cat("No ", name, " rows were produced for this dataset and specification.\n\n", sep = "")
    next
  }
  print(knitr::kable(summary_table, caption = heading))
  cat("\n\n")
}
```

## Plots

```{r plots, echo=FALSE, fig.width=7, fig.height=4}
if (length(results$plots) == 0L) {
  cat("No plots were produced for present specification variables.\n\n")
}
for (name in names(results$plots)) {
  cat("### ", name, "\n\n", sep = "")
  print(results$plots[[name]])
  cat("\n\n")
}
```

## Maps

```{r maps, echo=FALSE, fig.width=7, fig.height=5}
knitr::kable(results$map_inventory, caption = "Map inventory")
if (length(results$maps) == 0L) {
  cat("No maps were created.\n\n")
}
for (map_id in names(results$maps)) {
  cat("### ", map_id, "\n\n", sep = "")
  print(results$maps[[map_id]])
  cat("\n\n")
}
```
